---
title: "Codebase Starter Devcontainer"
author:
  - David Kane
  - Krrish Jayakumar
tutorial:
  id: "11-codebase-start"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "A tour of the devcontainer.json behind codespace-starter itself --- the custom PPBDS image, the VS Code extensions array, the settings that control auto-save, plots, and Git commits, and the postCreateCommand that keeps course packages fresh."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In `10-devcontainer` you examined a generic Python devcontainer and wrote a minimal R devcontainer from scratch. Now you will look at the devcontainer behind the Codespace you are working in right now. `codespace-starter` uses a custom image built specifically for this course and a detailed `settings` block that configures VS Code's behavior. Many of the behaviors you may have noticed --- auto-save, plots appearing inside the editor, the Git commit button working without an extra confirmation step --- are controlled here.

###

You do not need to create a new repo for this tutorial. All of the exercises below work directly inside the `codespace-starter` checkout that is already part of this Codespace.

## Codespace Starter Devcontainer
###

### Exercise 1

In the bash Terminal, run `cd /workspaces/codespace-starter`, then `cat .devcontainer/devcontainer.json`. Copy and paste the entire output.

```{r codespace-starter-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Your answer should look something like:

````
{
  "name": "PPBDS Codespace",
  "image": "ghcr.io/ppbds/devcontainer:1.0.1",
  "customizations": {
    "vscode": {
      "extensions": [
        "reditorsupport.r",
        "quarto.quarto",
        "PPBDS.vscode-r-tutorials",
        "ritwickdey.LiveServer",
        "tomoki1207.pdf",
        "mechatroner.rainbow-csv",
        // Python data science: language support + notebooks. Unlike the AI
        // assistant panels above, these are ordinary language tooling that
        // renders fine in the web Codespaces client.
        "ms-python.python",
        "ms-toolsai.jupyter"
      ],
      "settings": {
        "r.rterm.linux": "/usr/local/bin/arf",
        "r.plot.useHttpgd": true,
        "r.source.echo": true,
        "git.enableSmartCommit": true,
        "files.autoSave": "afterDelay",
        "editor.wordWrap": "on",
        "workbench.editor.enablePreview": false,
        "python.defaultInterpreterPath": "/opt/venv/bin/python"
      }
    }
  },
  "postCreateCommand": "R -q -e 'pak::pkg_install(c(\"PPBDS/tutorial.helpers\", \"PPBDS/vscode.tutorials\", \"PPBDS/misc.tutorials\", \"PPBDS/primer.tutorials\"), ask = FALSE)'"
}
````

Note that the exact `name` field, key order, and image version may differ slightly on your machine as the image is updated.

###

Standard JSON doesn't allow comments, but `devcontainer.json` uses **JSONC** (JSON with Comments), which does. The comments here explain *why* each setting exists --- useful for instructors and future maintainers.

### Exercise 2

Find the line beginning with `"image"`. Copy and paste it.

```{r codespace-starter-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"image": "ghcr.io/ppbds/devcontainer:1.0.1",
```

Note that the version number will change as the image is updated.

###

`codespace-starter` uses a custom PPBDS image **pinned** to a specific version, so every student gets the exact same setup regardless of when they open the Codespace. There is a single image, `ghcr.io/ppbds/devcontainer`, that serves the whole PPBDS organization --- students taking the course and developers working on PPBDS packages alike. It pre-installs everything the course needs: Quarto, `gh`, the `arf` R console, `httpgd`, `pak`, the four course tutorial packages (`tutorial.helpers`, `vscode.tutorials`, `misc.tutorials`, and `primer.tutorials`), R stacks for modeling, mapping, and presentation, a Python data-science stack, and four AI command-line assistants --- `claude`, `codex`, `agy`, and `aider`. The next tutorial, `12-docker`, looks inside the actual build file that produces this image.

### Exercise 3

Find the `"extensions"` array. Copy and paste the entire block, from the line containing `"extensions": [` through the closing `]`.

```{r codespace-starter-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

Your answer should look like:

```
"extensions": [
  "reditorsupport.r",
  "quarto.quarto",
  "PPBDS.vscode-r-tutorials",
  "ritwickdey.LiveServer",
  "tomoki1207.pdf",
  "mechatroner.rainbow-csv",
  // Python data science: language support + notebooks. Unlike the AI
  // assistant panels above, these are ordinary language tooling that
  // renders fine in the web Codespaces client.
  "ms-python.python",
  "ms-toolsai.jupyter"
],
```

###

Each extension serves a specific role: `reditorsupport.r` provides R language support, `quarto.quarto` enables Quarto document rendering, `PPBDS.vscode-r-tutorials` powers the tutorial panel you are using right now, `ritwickdey.LiveServer` lets you preview HTML files in the browser, `tomoki1207.pdf` adds a PDF viewer directly inside VS Code, `mechatroner.rainbow-csv` adds a CSV viewer with color-coded columns, and `ms-python.python` with `ms-toolsai.jupyter` add Python language support and Jupyter notebooks. Notice what is *absent*: there are no AI-assistant extensions. The AI tools ship as command-line programs instead (`claude`, `codex`, `agy`, `aider`), because the in-editor AI panels render poorly in the browser-based Codespaces client --- a CLI works on every device.

### Exercise 4

Find the line containing `"r.rterm.linux"`. Copy and paste it.

```{r codespace-starter-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"r.rterm.linux": "/usr/local/bin/arf",
```

###

`arf` is a fast, Rust-based replacement for the default R Terminal: it starts more quickly and handles pasted multi-line code more reliably. The path points at `/usr/local/bin/arf`, a stable symlink the image provides for exactly this purpose, so the setting keeps working even if the image's internals move.

### Exercise 5

Find the line containing `"r.plot.useHttpgd"`. Copy and paste it.

```{r codespace-starter-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"r.plot.useHttpgd": true,
```

###

By default, R opens plots in a separate window that would be invisible in a Codespace. Setting this to `true` sends plots to the `httpgd` tool instead, which streams them into a panel inside VS Code.

### Exercise 6

Find the line containing `"r.source.echo"`. Copy and paste it.

```{r codespace-starter-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"r.source.echo": true,
```

###

The "Run Source" button (the ▷ in the editor title bar) and its `Ctrl/Cmd + Shift + S` shortcut normally run a file silently, showing only the final output. This setting makes them run `source()` with `echo = TRUE` instead, so each line of code is printed as it executes --- when something goes wrong, you can see exactly which line produced the error.

### Exercise 7

Find the line containing `"git.enableSmartCommit"`. Copy and paste it.

```{r codespace-starter-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"git.enableSmartCommit": true,
```

###

Smart commit automatically includes all changed files in the commit when you click the Commit button in the Source Control sidebar --- you do not need to run `git add` separately. This is why clicking Commit in the `codespace-starter` Codespace works without an extra confirmation step, while the Jupyter Notebook template required one.

### Exercise 8

Find the line containing `"files.autoSave"`. Copy and paste it.

```{r codespace-starter-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"files.autoSave": "afterDelay",
```

###

With `"afterDelay"`, VS Code automatically saves open files a short time after you stop typing. This prevents the most common beginner mistake in coding environments: running code that still reflects the last-saved version rather than the current edits, because the file was never manually saved.

### Exercise 9

Find the line containing `"editor.wordWrap"`. Copy and paste it.

```{r codespace-starter-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"editor.wordWrap": "on",
```

###

Word wrap causes long lines to wrap to the next line visually rather than extending off the right edge of the screen. Without it, long comment blocks or URLs in code appear truncated, and students on small screens may not notice that content is hidden to the right.

### Exercise 10

Find the line containing `"workbench.editor.enablePreview"`. Copy and paste it.

```{r codespace-starter-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"workbench.editor.enablePreview": false,
```

###

By default, VS Code opens files in "preview" mode --- clicking a second file replaces the first tab, so beginners often think their file disappeared. Setting this to `false` means every file opens in a permanent tab.

### Exercise 11

Find the line beginning with `"python.defaultInterpreterPath"`. Copy and paste it.

```{r codespace-starter-11}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
"python.defaultInterpreterPath": "/opt/venv/bin/python",
```

###

The image ships a full Python data-science stack --- numpy, pandas, matplotlib, scikit-learn, and more --- in a virtual environment at `/opt/venv`. This setting points the Python extension at that environment, so `import pandas` works in a Jupyter notebook or a Quarto Python chunk with no setup. This is a recent addition: earlier versions of `codespace-starter` were R-only.

### Exercise 12

Find the line beginning with `"postCreateCommand"`. Copy and paste it.

```{r codespace-starter-12}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look like:

```
"postCreateCommand": "R -q -e 'pak::pkg_install(c(\"PPBDS/tutorial.helpers\", \"PPBDS/vscode.tutorials\", \"PPBDS/misc.tutorials\", \"PPBDS/primer.tutorials\"), ask = FALSE)'",
```

###

Like the Jupyter template's `updateContentCommand`, this runs automatically when the Codespace is created. It refreshes the four fast-moving course packages --- `tutorial.helpers`, `vscode.tutorials`, `misc.tutorials`, and `primer.tutorials` --- to their latest versions, so you always get the newest tutorials even though the image itself is pinned. The image bakes in working copies as a fallback, so a momentary GitHub outage never blocks the Codespace from opening.

## Summary
###

This tutorial toured the `devcontainer.json` behind `codespace-starter` itself: its pinned custom PPBDS image, its VS Code extensions array, the settings that control the R Terminal, plot rendering, Git commits, autosave, word wrap, tab behavior, and the Python interpreter, and the `postCreateCommand` that keeps the four fast-moving course packages current. The next tutorial, `12-docker`, goes one level deeper and reads through the Dockerfile that actually builds the `ghcr.io/ppbds/devcontainer` image this `devcontainer.json` points at.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
