---
title: "Models And Money"
author: David Kane
tutorial:
  id: "16-models"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "What is underneath a coding agent --- one HTTP request. Call OpenRouter directly from bash and from R with `ellmer`, compare what open weights and frontier models cost per million tokens, and decide whether to buy $10 of credits."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In `15-openrouter` you drove a model through `aider`. The agent did the talking. Here you will take the agent away and talk to the model yourself --- first with `curl`, then from R.

###

There are two reasons to bother. The first is that a coding agent is a thin wrapper around a single HTTP request, and seeing that request stops the whole business from looking like magic. The second is that most of the useful work a model does for a data scientist is not writing code at all. It is reading ten thousand rows of messy free text and telling you what is in them --- and for that you need to call a model from a script, in a loop, cheaply.

###

Cheaply is where this tutorial ends up. By the last exercise you will know what a model costs to the dollar, why some cost a hundred times less than others, and whether it is worth giving OpenRouter ten of yours.

### Exercise 1

Create and connect to a repo called `16-models`, then open it as your workspace.

Copy and paste the URL for your `16-models` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://github.com/ppbds-student/16-models
```

###

You need the OpenRouter key from `15-openrouter` for everything here. If you saved it as a Codespaces secret and this is a *new* Codespace, it is already waiting for you.

### Exercise 2

In the bash Terminal, check that your key is present:

<pre><code>echo ${OPENROUTER_API_KEY:0:8}
</code></pre>

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
$ echo ${OPENROUTER_API_KEY:0:8}
sk-or-v1
```

###

If that came back empty, your Codespaces secret has not reached this shell. Either you are in the same Codespace as last time --- secrets arrive when a Codespace is *created*, not when it restarts --- or the secret was never granted access to `PPBDS/codespace-starter`. Fix it at <https://github.com/settings/codespaces>, or just `export` the key again for this session as you did last time.

That distinction is worth holding onto, because it is a general property of container configuration rather than a quirk of GitHub: things wired in at creation time do not appear in a machine that already exists. It is the same reason a change to `devcontainer.json` does nothing until you rebuild.

## The Request Underneath
###

Every AI product you have ever used --- the chat window, the editor sidebar, `aider`, Claude Code --- ends up sending the same thing: a list of messages to an HTTP endpoint, with a key in the header. This section is that request, with nothing on top of it.

### Exercise 1

Ask OpenRouter how many models it offers. In the bash Terminal:

<pre><code>curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
</code></pre>

CP/CR.

```{r api-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
$ curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
341
```

###

Your number will differ from ours by a few dozen, because models arrive and leave constantly.

Notice what that request did *not* need: a key. The catalogue is public. Notice also what a router buys you, now that you can count it --- several hundred models from dozens of companies, reachable with one account, one key, and one bill. The alternative is an account with OpenAI, another with Anthropic, another with Google, another with DeepSeek, four keys to store, four invoices, and four slightly different APIs to write code against.

### Exercise 2

Now send an actual prompt. Replace the model with any free one from your `15-openrouter` list if this one is gone:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "In one sentence, what is a tibble?"}]}' \
  | python -m json.tool
</code></pre>

CP/CR.

```{r api-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 16)
```

###

<!-- VERIFY: response envelope field set. -->

```
{
    "id": "gen-1753632000-abc123",
    "model": "openai/gpt-oss-20b:free",
    "object": "chat.completion",
    "choices": [
        {
            "message": {
                "role": "assistant",
                "content": "A tibble is the tidyverse's version of a data frame, with stricter behavior and much nicer printing."
            },
            "finish_reason": "stop"
        }
    ],
    "usage": {
        "prompt_tokens": 17,
        "completion_tokens": 24,
        "total_tokens": 41
    }
}
```

###

That is the whole of it. A model name, a list of messages each tagged with a `role`, and a key in the header. Everything else --- the chat interfaces, the agents, the editor plugins --- is software wrapped around this one call.

Two fields repay a close look. `finish_reason` says `stop`, meaning the model finished its thought; when it says `length` instead, the model was cut off mid-sentence by a token limit, which is the actual explanation for most "the AI stopped in the middle" complaints. And `usage` is the meter: 17 tokens in, 24 out. That is what you are billed for, and it is why the next section is about tokens rather than about questions or minutes.

### Exercise 3

Prices are published in that same catalogue. Have AI write you a command that reads <https://openrouter.ai/api/v1/models> and prints the input and output price per million tokens for these three models:

<pre><code>qwen/qwen3-coder-30b-a3b-instruct
deepseek/deepseek-v3.2
anthropic/claude-opus-4.8
</code></pre>

Run it. CP/CR.

```{r api-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

```
    0.07     0.27  qwen/qwen3-coder-30b-a3b-instruct
    0.27     0.40  deepseek/deepseek-v3.2
    5.00    25.00  anthropic/claude-opus-4.8
```

###

The catalogue is machine-readable, which is the part that matters. Prices are not marketing copy on a pricing page you have to go read; they are data, in the same response your program already fetches, and you can sort by them.

Look at the spread. Output tokens on Opus cost ninety-three times what they cost on the Qwen coder model. Not ninety-three percent more --- ninety-three times. That ratio is the subject of the next section.

## Open Weights
###

Why should two things that both answer questions in English differ in price by two orders of magnitude? The answer is not mostly about quality. It is about who is allowed to sell them.

### Exercise 1

Ask the same question of a cheap open weights model and of an expensive frontier model, and compare.

First the cheap one:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "Write R code to count rows per cut in the diamonds data set."}]}' \
  | python -c "import sys, json; print(json.load(sys.stdin)['choices'][0]['message']['content'])"
</code></pre>

CP/CR.

```{r weights-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

```
library(tidyverse)

diamonds |>
  count(cut)
```

###

Piping the response through `python` to pull out `choices[0].message.content` is how you get the answer without the envelope around it --- useful the moment you want a model's output inside a script rather than on your screen.

That is a correct, idiomatic, three-line answer from a model that charged you nothing. Hold it next to what you would have paid a frontier model for the identical result.

### Exercise 2

Run the same request against `anthropic/claude-opus-4.8`.

This one is **not** free, and your account has no credits, so it will fail. Copy and paste the error.

```{r weights-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

<!-- VERIFY: exact 402 message text for a zero-credit account. -->

```
{
    "error": {
        "message": "Insufficient credits. Add more using https://openrouter.ai/settings/credits",
        "code": 402
    }
}
```

###

A `402` is the HTTP status code for *Payment Required* --- one of the rarest codes on the ordinary web, and routine on this one.

Now the reason for the price gap. **Open weights** models --- `gpt-oss`, DeepSeek, Qwen, Llama, Gemma --- are published as files anyone can download and run. **Closed** models --- Opus, GPT, Gemini --- exist only on their owner's servers. That difference has an economic consequence that is entirely predictable: when a dozen companies can serve the same Qwen weights, they compete on price and margins collapse toward the cost of the electricity. When only Anthropic can serve Opus, the price is whatever Anthropic says it is.

So you are not usually paying ninety-three times more for something ninety-three times better. You are paying for a monopoly on the good ones, and paying commodity rates for the rest.

### Exercise 3

Work out what a real session costs. A busy hour with a coding agent runs to roughly 500,000 input tokens --- agents resend the whole conversation with every message --- and 50,000 output tokens.

Using the prices you printed earlier, have AI compute the cost of that hour on `qwen/qwen3-coder-30b-a3b-instruct` and on `anthropic/claude-opus-4.8`. Copy and paste both figures.

```{r weights-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
qwen3-coder-30b:  0.5 * 0.07 + 0.05 * 0.27  = $0.049
claude-opus-4.8:  0.5 * 5.00 + 0.05 * 25.00 = $3.75
```

###

Five cents against three dollars seventy-five, for one hour. Over a semester of daily use that is the difference between a few dollars and a few hundred.

Which does not mean always take the cheap one. It means the choice is now a real decision with a number attached, and the sensible policy is the one professionals actually use: cheap models for the bulk of the work --- boilerplate, renaming, tests, the tenth variation on a plot --- and an expensive model for the genuinely hard problem you are stuck on. That is exactly what `aider`'s `/architect` mode automates, and now you can see why it is worth the trouble.

## Models From R
###

Everything so far has been about writing code. But the reason a data scientist cares about cheap models is different, and larger: a model is a function you can call on every row of a data set.

### Exercise 1

In the R Terminal, run `library(ellmer)`. CP/CR.

```{r ellmer-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- ellmer must be in the PPBDS/devcontainers image before this ships; until then
     students hit "there is no package called 'ellmer'". Verified against ellmer
     0.4.2: chat_openrouter(model = ...) constructs, reads OPENROUTER_API_KEY,
     returns an R6 Chat object. -->

```
> library(ellmer)
```

###

**[ellmer](https://ellmer.tidyverse.org/)** is the tidyverse's package for talking to language models from R. It covers the same ground `curl` just covered, with the HTTP hidden: one function per provider, a chat object you can keep talking to, and the same `OPENROUTER_API_KEY` environment variable you have been using all along.

It is not the only option --- **[httr2](https://httr2.r-lib.org/)** would let you build that `POST` request by hand, and Python's `openai` library is the equivalent on the other side --- but `ellmer` is the one that fits the way you already work in R.

### Exercise 2

Still in the R Terminal, create a chat and ask it something:

<pre><code>chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat("In one sentence, what is a tibble?")
</code></pre>

CP/CR.

```{r ellmer-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

```
> chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
> chat$chat("In one sentence, what is a tibble?")
A tibble is the tidyverse's version of a data frame, with stricter behavior
and much nicer printing.
```

###

`chat_openrouter()` found your key without being told where it was --- it reads `OPENROUTER_API_KEY` out of the environment, the same variable `aider` and `curl` used. One key, three tools.

Always name the model. `ellmer` has a default, and its documentation says plainly that the default changes; code that relies on it will one day start costing money it did not cost yesterday.

### Exercise 3

Now the thing you could not do by hand. In the R Terminal:

<pre><code>reviews <- c("the room was filthy", "staff went out of their way for us",
             "fine, but overpriced", "would absolutely stay again")

chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat(paste("Label each line positive, negative, or mixed. One label per line, nothing else:",
                paste(reviews, collapse = "\n")))
</code></pre>

CP/CR.

```{r ellmer-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

```
negative
positive
mixed
positive
```

###

This is the use case that makes cheap models matter. Four hotel reviews are a toy; four *hundred thousand* of them is a real data set, and there has never been a way to turn free text into a usable variable at that scale. Hand-coding was the only option, and it meant hiring people.

Run the arithmetic. Four hundred thousand short reviews is on the order of 20 million input tokens. On the Qwen coder model that is about **$1.40**. On Opus it is **$100**, for a labeling job where the cheap model is very nearly as good. Cost is not a footnote to that decision; it is the reason the analysis happens at all.

The catch worth naming: a model asked for four labels will occasionally return five, or add a sentence of commentary. Production code pins the output down --- `ellmer` will make a model return a specified structure rather than free prose --- and always validates what came back before trusting it.

## Paying
###

You have done everything in two tutorials without spending anything. This section is about whether to stop there.

### Exercise 1

In the bash Terminal, check what you have spent:

<pre><code>curl -s https://openrouter.ai/api/v1/key -H "Authorization: Bearer $OPENROUTER_API_KEY" | python -c "import sys, json; d = json.load(sys.stdin)['data']; print('spent so far:', d['usage']); print('free tier:', d['is_free_tier'])"
</code></pre>

CP/CR.

```{r paying-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
spent so far: 0
free tier: True
```

###

Zero, which is where this course expects you to stay. **Nothing in this course requires you to give anyone a credit card**, and a free account with 50 requests a day is enough for every tutorial we will ask you to do.

Pulling two fields out of that response rather than printing all fourteen is worth noticing as a habit. An API gives you everything it has; a useful program takes the two things it came for.

It is still worth knowing exactly what the ten dollars buys, so that the decision is yours rather than ours. Buying credits once --- the minimum is $10, and OpenRouter takes about 5.5% --- does three things permanently: the free-model ceiling rises from 50 requests a day to 1,000; every paid model becomes available at the per-token prices you printed earlier; and you can turn those privacy toggles back off, because paid endpoints do not require your prompts as payment.

That last one is the reason a professional buys credits even if she never touches an expensive model. Ten dollars is what it costs to stop paying with your data. Whether that trade is worth it to you is a judgment, and you now have every number you need to make it.

### Exercise 2

Commit and push your work.

In the bash Terminal, run `git log --oneline`. CP/CR.

```{r paying-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
$ git log --oneline
7d21ab8 (HEAD -> main, origin/main) Add model comparison work
```

###

Nothing you committed contains your key --- worth confirming rather than assuming. The key lived in an environment variable and in GitHub's encrypted secret store, and never in a file. That is the arrangement to reproduce on every project you ever work on.

## Summary
###

You took the agent away and found one HTTP request underneath it: a model name, a list of messages, and a key in the header. Everything built on top of a language model is built on that.

###

You saw the price catalogue as data rather than marketing, and found a ninety-three-fold spread between an open weights model and a frontier one. The explanation is competition, not quality: published weights can be served by anyone, so their price falls to the cost of running them, while a closed model is priced by the only company allowed to sell it. That is why the sensible working habit is a cheap model for most of the work and an expensive one for the problem you are actually stuck on.

###

You called a model from R with **ellmer**, using the same key, and used it for the job that has no other solution --- turning free text into a variable. And you finished where you started, at zero dollars, now knowing precisely what the ten-dollar decision would buy you.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
