---
title: "Models And Money"
author: David Kane
tutorial:
  id: "18-models"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "What is underneath a coding agent --- one HTTP request. Call OpenRouter directly from bash and from R with `ellmer`, compare what open weights and frontier models cost per million tokens, and decide whether to buy $10 of credits."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In the OpenRouter tutorial, `aider` talked to the model for you. Now you will take `aider` away and talk to the model yourself --- first from the bash Terminal, then from R.

###

Two reasons this is worth doing.

**It stops being magic.** Underneath every AI tool is one ordinary web request. Once you have sent that request by hand, the whole thing is much less mysterious.

**Most useful AI work is not writing code.** It is reading ten thousand rows of messy text and telling you what is in them. For that you need to call a model from a script, over and over, cheaply.

###

By the last exercise you will know what a model costs, why some cost a hundred times less than others, and whether $10 is worth spending.

### Exercise 1

Create and connect to a repo called `models-and-money`, then open it as your workspace.

Copy and paste the URL for your `models-and-money` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://github.com/ppbds-student/models-and-money
```

###

You need the OpenRouter key from the OpenRouter tutorial for everything here. If you saved it as a Codespaces secret and this is a *new* Codespace, it is already waiting for you.

### Exercise 2

In the bash Terminal, check that your key is present:

<pre><code>echo ${OPENROUTER_API_KEY:0:8}
</code></pre>

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
models-and-money $ echo ${OPENROUTER_API_KEY:0:8}
sk-or-v1
```

###

If that came back empty, the secret has not reached this Terminal. Either this is the same Codespace as last time --- secrets only arrive when a Codespace is *created* --- or the secret was never given access to `PPBDS/codespace-starter`. Fix it at <https://github.com/settings/codespaces>, or just `export` the key again like last time.

That rule is worth remembering: settings applied when a machine is built do not appear on a machine that already exists. It is the same reason editing `devcontainer.json` does nothing until you rebuild.

## The Request Underneath
###

Every AI product you have used --- the chat window, the editor sidebar, `aider`, Claude Code --- sends the same thing underneath: a list of messages to a web address, with a key attached. This section is that request, with nothing built on top of it.

### Exercise 1

Ask OpenRouter how many models it offers. In the bash Terminal:

<pre><code>curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
</code></pre>

CP/CR.

```{r api-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
models-and-money $ curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
341
```

###

Your number will be a bit different from ours. Models arrive and leave all the time.

Notice what that request did *not* need: a key. The catalogue is public to anyone.

And notice what a router gets you, now that you can count it. Several hundred models, from dozens of companies, through one account and one key. Without a router you would need an account with OpenAI, another with Anthropic, another with Google, another with DeepSeek --- four keys to keep track of and four different ways of asking.

### Exercise 2

Now send an actual prompt. Replace the model with any free one from the list you made in the OpenRouter tutorial if this one is gone:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "In one sentence, what is a tibble?"}]}' \
  | python -m json.tool
</code></pre>

CP/CR.

```{r api-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Ours, with the model's chain of thought cut short to fit:

```
{
    "id": "gen-1785184071-Ek9msxDQCT6FtDLjnFt7",
    "object": "chat.completion",
    "created": 1785184071,
    "model": "openai/gpt-oss-20b:free",
    "provider": "Darkbloom",
    "choices": [
        {
            "index": 0,
            "finish_reason": "stop",
            "native_finish_reason": "stop",
            "message": {
                "role": "assistant",
                "content": "A tibble is a modern, tidyverse-centric re-implementation of R's data frame that retains tabular structure but offers cleaner printing, no automatic coercion to factors, and a consistent, column-wise subsetting interface.",
                "reasoning": "We need to answer: \"In one sentence, what is a tibble?\" Probably a brief definition ...",
                "refusal": null
            }
        }
    ],
    "usage": {
        "prompt_tokens": 77,
        "completion_tokens": 177,
        "total_tokens": 254,
        "cost": 0,
        "cost_details": {
            "upstream_inference_cost": 1.35065e-05
        },
        "completion_tokens_details": {
            "reasoning_tokens": 120
        }
    }
}
```

###

That is the whole thing: a model name, a list of messages, and a key. Every chat window and coding assistant is software wrapped around this one request.

Four fields in the reply are worth a look.

**`provider`** is the company that actually ran the model. You asked for `openai/gpt-oss-20b:free`, and OpenRouter chose a company called Darkbloom to do it. That is the "routing" part: several companies host the same model, and the service picked one. Yours will probably name a different one.

**`reasoning`** is the model thinking out loud before answering. Look at what that costs: **120 of the 177 output tokens were thinking, not answer.** You pay for the thinking too, and it is usually the bigger half.

**`cost`** is `0`, as promised. But `upstream_inference_cost` shows the real cost was about **$0.0000135** --- and somebody paid it. Free means someone else is paying because they want your prompts.

**`finish_reason`** says `stop`, meaning the model finished its thought. If it says `length`, the model was cut off partway. That is the real reason behind most "the AI stopped in the middle" complaints.

### Exercise 3

Prices are published in that same catalogue. Have AI write you a command that reads <https://openrouter.ai/api/v1/models> and prints the input and output price per million tokens for these three models:

<pre><code>qwen/qwen3-coder-30b-a3b-instruct
deepseek/deepseek-v3.2
anthropic/claude-opus-4.8
</code></pre>

Run it. CP/CR.

```{r api-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

```
    0.07     0.27  qwen/qwen3-coder-30b-a3b-instruct
    0.27     0.40  deepseek/deepseek-v3.2
    5.00    25.00  anthropic/claude-opus-4.8
```

###

The prices are **data**, not marketing text on a web page. They arrive in the same response your program already asked for, so your program can sort by them.

Now look at the spread. Output on Opus costs ninety-three times what it costs on the Qwen model. Not ninety-three percent more. Ninety-three *times*. The next section explains why.

## Open Weights
###

Why should two things that both answer questions in English differ in price by a hundred times? The answer is mostly not about quality. It is about who is allowed to sell them.

### Exercise 1

Ask the same question of a cheap open weights model and of an expensive frontier model, and compare.

First the cheap one:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "Write R code to count rows per cut in the diamonds data set."}]}' \
  | python -c "import sys, json; print(json.load(sys.stdin)['choices'][0]['message']['content'])"
</code></pre>

CP/CR.

```{r weights-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 20)
```

###

Ours, abridged --- it ran to 546 tokens and offered three different solutions:

```
Below is a compact **R** script that will give you the number of observations
for each level of the `cut` variable in the built-in `diamonds gegeben` data
set from the **ggplot2** package.

## 2a. Plain R base solution (very concise)
cut_counts_base <- table(diamonds$cut)

## 2b. dplyr / tidyverse solution
cut_counts_tidy <- diamonds %>%
  count(cut)

### What you'll see

   Fair Good Ideal Premium       Very Good
    475    10105    12558    68146          46978
```

###

Look at those last four lines, then look at the truth. The real counts are Fair 1610, Good 4906, Very Good 12082, Premium 13791, Ideal 21551.

**Every number the model printed is made up.** `diamonds` has 53,940 rows. The model claims 68,146 Premium diamonds on their own, and its five numbers add up to more than twice the whole data set.

The code is fine --- `table(diamonds$cut)` and `count(cut)` both work. The problem is that the model never ran anything. It cannot. So when it wrote a heading saying "What you'll see," it wrote text that *looks* like output. There is also a German word stuck in the first sentence, which tells you how carefully this was put together.

Here is the habit to take from this tutorial, and it matters more than anything else in it:

> **You can check a model's code by running it. You cannot check a model's claim about what the code prints.**

So run it yourself. Every time. A cheap model is a fine way to get `count(cut)` written for you. It is no way at all to find out how many Ideal diamonds there are.

### Exercise 2

Run the same request against `anthropic/claude-opus-4.8`, piping through `python -m json.tool` this time rather than pulling out the content.

This one is **not** free, and your account has no credits, so it will fail. Copy and paste the error.

```{r weights-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```
{
    "error": {
        "message": "This request requires more credits, or fewer max_tokens. You requested up to 65536 tokens, but can only afford 1600. To increase, visit https://openrouter.ai/settings/credits and upgrade to a paid account",
        "code": 402,
        "metadata": {
            "provider_name": null,
            "previous_errors": [ ... ]
        }
    }
}
```

###

`402` means *Payment Required*. You will almost never see it on the ordinary web. Here it is routine.

Read what it actually says, though, because it is not quite "you have no money." It says you cannot afford the *worst case*. The model is allowed to write up to 65,536 tokens, so OpenRouter sets aside the cost of all 65,536 before starting, and you can only cover 1,600. Ask for a shorter answer and the same request would work.

Services that charge by use almost always reserve the maximum like this. Worth knowing before it surprises you.

`previous_errors` has four copies of the same complaint. OpenRouter tried four different companies for Opus, and all four said no. The routing worked fine. There was just nowhere to route to.

Now the reason for the price gap.

**Open weights** models --- `gpt-oss`, DeepSeek, Qwen, Llama, Gemma --- are published as files that anyone can download and run. **Closed** models --- Opus, GPT, Gemini --- live only on their owner's computers.

That difference decides the price. When a dozen companies can all run the same Qwen model, they compete, and the price falls to roughly what the electricity costs. When only Anthropic is allowed to run Opus, the price is whatever Anthropic says it is.

So you are usually not paying ninety-three times more for something ninety-three times better. You are paying for the fact that only one company sells it.

### Exercise 3

Work out what a real hour of work costs. A busy hour with a coding assistant uses roughly 500,000 input tokens (remember, the whole conversation is resent every message) and 50,000 output tokens.

Using the prices you printed earlier, have AI work out the cost of that hour on `qwen/qwen3-coder-30b-a3b-instruct` and on `anthropic/claude-opus-4.8`. Copy and paste both figures.

```{r weights-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
qwen3-coder-30b:  0.5 * 0.07 + 0.05 * 0.27  = $0.049
claude-opus-4.8:  0.5 * 5.00 + 0.05 * 25.00 = $3.75
```

###

Five cents against three dollars and seventy-five. For one hour. Over a semester of daily use, that is the difference between a few dollars and a few hundred.

This does not mean always pick the cheap one. It means you now have a real number to decide with. What professionals actually do is use a cheap model for most of the work --- renaming things, writing tests, the tenth version of a plot --- and an expensive model for the problem they are genuinely stuck on.

That is exactly what `aider`'s `/architect` mode does for you, and now you can see why it is worth the trouble.

## Models From R
###

Everything so far has been about writing code. But the reason a data scientist cares about cheap models is bigger than that: **a model is something you can run on every row of a data set.**

### Exercise 1

In the R Terminal, run `library(ellmer)`. CP/CR.

```{r ellmer-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- ellmer must be in the PPBDS/devcontainers image before this ships; until then
     students hit "there is no package called 'ellmer'". Verified against ellmer
     0.4.2: chat_openrouter(model = ...) constructs, reads OPENROUTER_API_KEY,
     returns an R6 Chat object. -->

```
> library(ellmer)
```

###

**[ellmer](https://ellmer.tidyverse.org/)** is the tidyverse package for talking to AI models from R. It does the same job as the `curl` commands above, but hides the web request. You get one function per service, a chat object you can keep talking to, and it uses the same `OPENROUTER_API_KEY` you have been using all along.

It is not the only choice. **[httr2](https://httr2.r-lib.org/)** would let you build the request by hand, and Python has the `openai` library. But `ellmer` fits the way you already work in R.

### Exercise 2

Still in the R Terminal, create a chat and ask it something:

<pre><code>chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat("In one sentence, what is a tibble?")
</code></pre>

CP/CR.

```{r ellmer-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

```
> chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
> chat$chat("In one sentence, what is a tibble?")
User asks: "In one sentence, what is a tibble?" Provide one sentence
definition. Something like "In R, a tibble is a modern re-implementation of
data frames that is part of the tidyverse." Should answer in one sentence.
Alright.A **tibble** is a tidy-verse-style, modern reimplementation of R's
data frame that stores tabular data as a list of consistent-typed columns
with a clean, concise printing, and features straightforward subsetting.
```

###

`chat_openrouter()` found your key on its own. It read `OPENROUTER_API_KEY`, the same variable `aider` and `curl` used. One key, three tools.

Now look at what came back. The first few lines are not the answer. They are the model's *reasoning* --- the same `reasoning` field you saw in the JSON earlier, which `ellmer` prints too. You can see exactly where thinking stops and answering starts: `Alright.A **tibble** is`, with no space, because two separate pieces got stuck together.

Your output will not match ours word for word, and may be messier than ours. Free models are run by whichever company is giving them away that day, and those are not always the ones being careful.

Always name the model. `ellmer` has a default, and its own documentation warns that the default changes. Code that relies on it may start costing money one day without you touching it.

### Exercise 3

Now the thing you could not do by hand. In the R Terminal:

<pre><code>reviews <- c("the room was filthy", "staff went out of their way for us",
             "fine, but overpriced", "would absolutely stay again")

chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat(paste("Label each line positive, negative, or mixed. One label per line, nothing else:",
                paste(reviews, collapse = "\n")))
</code></pre>

CP/CR.

```{r ellmer-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

Ours, with the reasoning trimmed --- the last four lines are what it offered as the answer:

```
1. "the room was filthy" -> Negative (clear negative).
2. staff ... it's a compliment about staff, positive.
3. "fine, but overpriced" -> Mixed?  ...
We need just one per line, no additional text.
negative
positive
mixed
 perhatian positif
```

###

Three labels are right. The fourth is `perhatian positif` --- Indonesian for "positive attention." We asked for one label per line and nothing else. We got a paragraph of thinking, three correct labels, and then a word in the wrong language.

This one result shows both why cheap models matter and why you cannot trust them blindly.

Four hotel reviews is a toy. Four *hundred thousand* is a real data set, and until recently the only way to turn that much writing into a usable column was to pay people to read it. Four hundred thousand short reviews is roughly 20 million tokens: about **$1.40** on the Qwen model, about **$100** on Opus.

But a column containing `perhatian positif` is not data, it is a bug. Two habits fix this, and professionals use both:

1. **Ask for a fixed shape.** `ellmer` can require the model to return one of exactly three labels instead of free writing. A wrong answer then becomes an error you can catch.
2. **Check anyway.** Count that you got back as many labels as you sent rows, and that every label is one you expected.

Neither is optional once the data is big. With four reviews you can spot the bad one. With four hundred thousand you cannot.

## Paying
###

You have done everything in two tutorials without spending anything. This section is about whether to stop there.

### Exercise 1

In the bash Terminal, check what you have spent:

<pre><code>curl -s https://openrouter.ai/api/v1/key -H "Authorization: Bearer $OPENROUTER_API_KEY" | python -c "import sys, json; d = json.load(sys.stdin)['data']; print('spent so far:', d['usage']); print('free tier:', d['is_free_tier'])"
</code></pre>

CP/CR.

```{r paying-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
spent so far: 0
free tier: True
```

###

Zero, and that is where this course expects you to stay. **Nothing in this course needs a credit card.** A free account with 50 requests a day covers every tutorial we will ask you to do.

Notice the habit in that command, too: the reply had fourteen fields and we asked for two. An API hands you everything it has. A good program takes only what it came for.

Still, you should know what $10 would buy, so the choice is yours rather than ours. Buying credits once --- the minimum is $10, and OpenRouter keeps about 5.5% --- changes three things for good:

- your free-model limit goes from 50 requests a day to 1,000
- every paid model becomes available at the prices you printed earlier
- you can turn those privacy switches back **off**, because paid models do not want your prompts as payment

That last one is why a professional might buy credits even if they never touch an expensive model. Ten dollars is the price of no longer paying with your data. Whether that is worth it is your call, and you now have every number you need to make it.

### Exercise 2

Commit and push your work.

In the bash Terminal, run `git log --oneline`. CP/CR.

```{r paying-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
models-and-money $ git log --oneline
7d21ab8 (HEAD -> main, origin/main) Add model comparison work
```

###

Nothing you committed contains your key. That is worth checking rather than assuming. The key lived in an environment variable and in GitHub's encrypted store, and never in a file. Copy that arrangement on every project you ever work on.

## Summary
###

You took the assistant away and found one ordinary web request underneath: a model name, a list of messages, and a key. Everything built on top of an AI model is built on that.

###

You saw prices as data rather than marketing, and found a ninety-three-fold gap between a cheap model and an expensive one. The reason is competition, not quality. Anyone can run a published model, so its price falls to the cost of running it. Only one company can run a closed model, so it charges what it likes.

That is why the sensible habit is a cheap model for most work and an expensive one for the problem you are truly stuck on.

###

You called a model from R with **ellmer** and used it for the job nothing else can do: turning ordinary writing into a column of data. You also saw it get one of four labels wrong, which is why checking the output is part of the job.

And you finished where you started --- at zero dollars, now knowing exactly what $10 would buy.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
