---
title: "Models And Money"
author: David Kane
tutorial:
  id: "18-models"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "What is underneath a coding agent --- one HTTP request. Buy $10 of OpenRouter credits with a real credit card, call the API from bash and from R with `ellmer`, and find out how much of the hundred-fold gap between a free model and a frontier one is quality and how much is ownership."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In the OpenRouter tutorial, `aider` talked to the model for you. Now you will take `aider` away and talk to the model yourself --- first from the bash Terminal, then from R.

###

Two reasons this is worth doing.

**It stops being magic.** Underneath every AI tool is one ordinary web request. Once you have sent that request by hand, the whole thing is much less mysterious.

**Most useful AI work is not writing code.** It is reading ten thousand rows of messy text and telling you what is in them. For that you need to call a model from a script, over and over, cheaply.

###

By the last exercise you will know what a model costs, why some cost a hundred times less than others, and how much of that gap is quality rather than ownership.

###

One thing to settle before you start, because this tutorial is different from every other one in the course.

**You need a real credit card.** Partway through you will buy $10 of credits on OpenRouter. The OpenRouter tutorial was careful to keep you at zero and promised you would never need a card for it, and that promise held. This tutorial is the exception, and here is why it has to be.

A free account cannot run a paid model at all. Not slowly, not in a limited way --- the request is refused. So on a free account there is no way to answer the question this tutorial exists to ask: is the expensive model actually better, and by how much? You would be reading our word for it instead of finding out.

###

Be clear about what the $10 is, because it is easy to imagine something worse than it is.

It is a **one-off purchase of credits**. Not a subscription. Nothing renews, nothing needs cancelling, and there is no monthly charge to forget about. You buy $10 once, spend it slowly, and when it eventually runs out you decide again from scratch.

The work in this tutorial costs a few cents. Almost all of the $10 is still sitting there when you finish, and it buys you two things that outlast today: your free-model limit goes from 50 requests a day to 1,000, and you can turn back off the data-training switches you had to turn on last time.

###

If you would rather not spend it, decide that now rather than halfway through. Talk to your instructor about running this tutorial from a shared account or watching it as a demonstration. Do not start it and stop at the paywall --- half of this tutorial is the half after the purchase.

### Exercise 1

You should be doing this tutorial in a repo named `models-and-money`. If you are not, create one and connect to it, as you learned in earlier tutorials.

Copy and paste the URL for your `models-and-money` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://github.com/ppbds-student/models-and-money
```

###

You need the OpenRouter key from the OpenRouter tutorial for everything here. If you saved it as a Codespaces secret and this is a *new* Codespace, it is already waiting for you.

### Exercise 2

In the bash Terminal, check that your key is present:

<pre><code>echo ${OPENROUTER_API_KEY:0:8}
</code></pre>

CP/CR.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
models-and-money $ echo ${OPENROUTER_API_KEY:0:8}
sk-or-v1
```

###

If that came back empty, the secret has not reached this terminal. Either this is the same Codespace as last time --- secrets only arrive when a Codespace is *created* --- or the secret was never given access to `PPBDS/codespace-starter`. Fix it at <https://github.com/settings/codespaces>, or just `export` the key again like last time.

That rule is worth remembering, and by now you have met it three times: settings applied when a machine is built do not appear on a machine that already exists. It is why editing `devcontainer.json` does nothing until you rebuild, and why `COOLEST_TEACHER` stayed empty in the Dotfiles tutorial until you opened a new Codespace.

## The Request Underneath
###

Every AI product you have used --- the chat window, the editor sidebar, `aider`, Claude Code --- sends the same thing underneath: a list of messages to a web address, with a key attached. This section is that request, with nothing built on top of it.

### Exercise 1

Ask OpenRouter how many models it offers. In the bash Terminal:

<pre><code>curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
</code></pre>

CP/CR.

```{r api-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
models-and-money $ curl -s https://openrouter.ai/api/v1/models | python -c "import sys, json; print(len(json.load(sys.stdin)['data']))"
341
```

###

Your number will be a bit different from ours. Models arrive and leave all the time.

Notice what that request did *not* need: a key. The catalogue is public to anyone.

And notice what a router gets you, now that you can count it. Several hundred models, from dozens of companies, through one account and one key. Without a router you would need an account with OpenAI, another with Anthropic, another with Google, another with DeepSeek --- four keys to keep track of and four different ways of asking.

### Exercise 2

Now send an actual prompt. Replace the model with any free one from the list you made in the OpenRouter tutorial if this one is gone:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "In one sentence, what is a tibble?"}]}' \
  | python -m json.tool
</code></pre>

CP/CR.

```{r api-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

Ours, with the model's chain of thought cut short to fit:

```
{
    "id": "gen-1785184071-Ek9msxDQCT6FtDLjnFt7",
    "object": "chat.completion",
    "created": 1785184071,
    "model": "openai/gpt-oss-20b:free",
    "provider": "Darkbloom",
    "choices": [
        {
            "index": 0,
            "finish_reason": "stop",
            "native_finish_reason": "stop",
            "message": {
                "role": "assistant",
                "content": "A tibble is a modern, tidyverse-centric re-implementation of R's data frame that retains tabular structure but offers cleaner printing, no automatic coercion to factors, and a consistent, column-wise subsetting interface.",
                "reasoning": "We need to answer: \"In one sentence, what is a tibble?\" Probably a brief definition ...",
                "refusal": null
            }
        }
    ],
    "usage": {
        "prompt_tokens": 77,
        "completion_tokens": 177,
        "total_tokens": 254,
        "cost": 0,
        "cost_details": {
            "upstream_inference_cost": 1.35065e-05
        },
        "completion_tokens_details": {
            "reasoning_tokens": 120
        }
    }
}
```

###

That is the whole thing: a model name, a list of messages, and a key. Every chat window and coding assistant is software wrapped around this one request.

Four fields in the reply are worth a look.

**`provider`** is the company that actually ran the model. You asked for `openai/gpt-oss-20b:free`, and OpenRouter chose a company called Darkbloom to do it. That is the "routing" part: several companies host the same model, and the service picked one. Yours will probably name a different one.

**`reasoning`** is the model thinking out loud before answering. Look at what that costs: **120 of the 177 output tokens were thinking, not answer.** You pay for the thinking too, and it is usually the bigger half.

**`cost`** is `0`, as promised. But `upstream_inference_cost` shows the real cost was about **$0.0000135** --- and somebody paid it. Free means someone else is paying because they want your prompts.

**`finish_reason`** says `stop`, meaning the model finished its thought. If it says `length`, the model was cut off partway. That is the real reason behind most "the AI stopped in the middle" complaints.

### Exercise 3

Prices are published in that same catalogue. Have AI write you a command that reads <https://openrouter.ai/api/v1/models> and prints the input and output price per million tokens for these three models:

<pre><code>qwen/qwen3-coder-30b-a3b-instruct
deepseek/deepseek-v3.2
anthropic/claude-opus-4.8
</code></pre>

Run it. CP/CR.

```{r api-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

```
    0.07     0.27  qwen/qwen3-coder-30b-a3b-instruct
    0.27     0.40  deepseek/deepseek-v3.2
    5.00    25.00  anthropic/claude-opus-4.8
```

###

The prices are **data**, not marketing text on a web page. They arrive in the same response your program already asked for, so your program can sort by them.

Now look at the spread. Output on Opus costs ninety-three times what it costs on the Qwen model. Not ninety-three percent more. Ninety-three *times*. The next section explains why.

## Open Weights
###

Why should two things that both answer questions in English differ in price by a hundred times? The answer is mostly not about quality. It is about who is allowed to sell them.

### Exercise 1

Ask the same question of a cheap open weights model and of an expensive frontier model, and compare.

First the cheap one:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "openai/gpt-oss-20b:free", "messages": [{"role": "user", "content": "Write R code to count rows per cut in the diamonds data set."}]}' \
  | python -c "import sys, json; print(json.load(sys.stdin)['choices'][0]['message']['content'])"
</code></pre>

CP/CR.

```{r weights-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 20)
```

###

Ours, abridged --- it ran to 546 tokens and offered three different solutions:

```
Below is a compact **R** script that will give you the number of observations
for each level of the `cut` variable in the built-in `diamonds gegeben` data
set from the **ggplot2** package.

## 2a. Plain R base solution (very concise)
cut_counts_base <- table(diamonds$cut)

## 2b. dplyr / tidyverse solution
cut_counts_tidy <- diamonds %>%
  count(cut)

### What you'll see

   Fair Good Ideal Premium       Very Good
    475    10105    12558    68146          46978
```

###

Look at those last four lines, then look at the truth. The real counts are Fair 1610, Good 4906, Very Good 12082, Premium 13791, Ideal 21551.

**Every number the model printed is made up.** `diamonds` has 53,940 rows. The model claims 68,146 Premium diamonds on their own, and its five numbers add up to more than twice the whole data set.

The code is fine --- `table(diamonds$cut)` and `count(cut)` both work. The problem is that the model never ran anything. It cannot. So when it wrote a heading saying "What you'll see," it wrote text that *looks* like output. There is also a German word stuck in the first sentence, which tells you how carefully this was put together.

Here is the habit to take from this tutorial, and it matters more than anything else in it:

> **You can check a model's code by running it. You cannot check a model's claim about what the code prints.**

So run it yourself. Every time. A cheap model is a fine way to get `count(cut)` written for you. It is no way at all to find out how many Ideal diamonds there are.

### Exercise 2

Run the same request against `anthropic/claude-opus-4.8`, piping through `python -m json.tool` this time rather than pulling out the content.

This one is **not** free, and your account has no credits, so it will fail. Copy and paste the error.

```{r weights-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```
{
    "error": {
        "message": "This request requires more credits, or fewer max_tokens. You requested up to 65536 tokens, but can only afford 1600. To increase, visit https://openrouter.ai/settings/credits and upgrade to a paid account",
        "code": 402,
        "metadata": {
            "provider_name": null,
            "previous_errors": [ ... ]
        }
    }
}
```

###

`402` means *Payment Required*. You will almost never see it on the ordinary web. Here it is routine.

Read what it actually says, though, because it is not quite "you have no money." It says you cannot afford the *worst case*. The model is allowed to write up to 65,536 tokens, so OpenRouter sets aside the cost of all 65,536 before starting, and you can only cover 1,600. Ask for a shorter answer and the same request would work.

Services that charge by use almost always reserve the maximum like this. Worth knowing before it surprises you.

`previous_errors` has four copies of the same complaint. OpenRouter tried four different companies for Opus, and all four said no. The routing worked fine. There was just nowhere to route to.

Now the reason for the price gap.

**Open weights** models --- `gpt-oss`, DeepSeek, Qwen, Llama, Gemma --- are published as files that anyone can download and run. **Closed** models --- Opus, GPT, Gemini --- live only on their owner's computers.

That difference decides the price. When a dozen companies can all run the same Qwen model, they compete, and the price falls to roughly what the electricity costs. When only Anthropic is allowed to run Opus, the price is whatever Anthropic says it is.

So you are usually not paying ninety-three times more for something ninety-three times better. You are paying for the fact that only one company sells it.

### Exercise 3

Now fix the `402`. This is the credit card exercise.

Go to <https://openrouter.ai/settings/credits> and buy **$10** of credits. Ten dollars is the minimum, and it is far more than this tutorial will spend. Do not buy more.

When the purchase has gone through, check it from the bash Terminal:

<pre><code>curl -s https://openrouter.ai/api/v1/key -H "Authorization: Bearer $OPENROUTER_API_KEY" | python -c "import sys, json; d = json.load(sys.stdin)['data']; print('spent so far:', d['usage']); print('free tier:', d['is_free_tier'])"
</code></pre>

CP/CR.

```{r weights-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
spent so far: 0
free tier: False
```

###

`is_free_tier` has flipped to `False`. That is the whole change, and it is the switch every paid model checks before it agrees to run.

Look at the other number, though. You have spent **nothing**. Buying credits is not spending them --- it is putting money in an account you now draw down a fraction of a cent at a time. `usage` will still be a very small number when you finish this tutorial.

OpenRouter keeps about 5.5% of what you put in, so your $10 buys roughly $9.45 of model time. That is the fee for having one account and one key instead of five.

###

Two things you can now do, neither of which is on today's list, both worth knowing.

Your daily free-model limit went from 50 requests to 1,000. Everything you did in the OpenRouter tutorial on free models just got twenty times more room.

And you can go back to <https://openrouter.ai/settings/privacy> and turn **off** the two data-training switches you turned on last time. Paid models do not want your prompts as payment, so you no longer have to offer them. For many professionals that alone is what the $10 is for.

### Exercise 4

Now run the request that was refused. Same model as Exercise 2, and this time pull out the answer and the cost rather than the whole reply:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "anthropic/claude-opus-4.8", "max_tokens": 400, "messages": [{"role": "user", "content": "Write R code to count rows per cut in the diamonds data set."}]}' \
  | python -c "import sys, json; d = json.load(sys.stdin); print(d['choices'][0]['message']['content']); print('--- cost:', d['usage']['cost'])"
</code></pre>

CP/CR.

```{r weights-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 16)
```

###

Ours:

```
library(ggplot2)
library(dplyr)

diamonds |>
  count(cut)

# Base R alternative:
# table(diamonds$cut)

--- cost: 0.0057375
```

###

Two things to notice, and the second one matters more.

**It cost half a cent.** Not half a cent per hour --- half a cent for that one question. Your $10 buys somewhere around seventeen hundred of them.

**It did not tell you what the answer is.** Go back and look at what the free model did with this identical prompt in Exercise 1: it invented a table of counts, under a heading saying "What you'll see," and every number in it was wrong. Opus wrote the code and stopped. It cannot run R either, and it did not pretend otherwise.

That is a real quality difference, and it is not the one people expect. The expensive model is not better at *knowing* how many Ideal diamonds there are. It is better at **not making something up when it does not know**. On a task where being confidently wrong is expensive, that is most of what you are buying.

###

One more detail in that command. We added `"max_tokens": 400`.

That is Exercise 2's error message put to work. OpenRouter reserves the cost of the worst case before it starts, so a request that allows 65,536 tokens of output needs you to be able to afford 65,536 tokens of output. Capping it means you reserve less, and it also means a model cannot run away and write you three thousand words you did not want.

Set `max_tokens` on anything you run in a loop. It is the cheapest insurance there is.

### Exercise 5

Work out what a real hour of work costs. A busy hour with a coding assistant uses roughly 500,000 input tokens (remember, the whole conversation is resent every message) and 50,000 output tokens.

Using the prices you printed earlier, have AI work out the cost of that hour on `qwen/qwen3-coder-30b-a3b-instruct` and on `anthropic/claude-opus-4.8`. Copy and paste both figures.

```{r weights-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
qwen3-coder-30b:  0.5 * 0.07 + 0.05 * 0.27  = $0.049
claude-opus-4.8:  0.5 * 5.00 + 0.05 * 25.00 = $3.75
```

###

Five cents against three dollars and seventy-five. For one hour. Over a semester of daily use, that is the difference between a few dollars and a few hundred.

This does not mean always pick the cheap one. It means you now have a real number to decide with. What professionals actually do is use a cheap model for most of the work --- renaming things, writing tests, the tenth version of a plot --- and an expensive model for the problem they are genuinely stuck on.

That is exactly what `aider`'s `/architect` mode does for you, and now you can see why it is worth the trouble.

## Cheap Models
###

"Cheap" so far has meant one model, `qwen/qwen3-coder-30b-a3b-instruct`, because we picked it for you. Now find the cheap end of the market yourself, and then see what you actually give up by working there.

### Exercise 1

Have AI write you a command that reads <https://openrouter.ai/api/v1/models> and prints the ten **cheapest models that are not free**, sorted by output price, showing input and output price per million tokens.

Run it. CP/CR.

```{r cheap-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 14)
```

###

Yours will differ from ours --- prices move and models come and go:

```
    0.01     0.03  meta-llama/llama-3.2-1b-instruct
    0.02     0.04  google/gemma-4-4b-it
    0.02     0.05  qwen/qwen3-1.7b-instruct
    0.03     0.09  mistralai/ministral-3b
    0.04     0.14  nvidia/nemotron-3-nano-30b-a3b
    0.05     0.18  inclusionai/ling-3.0-flash
    0.06     0.22  openai/gpt-oss-20b
    0.07     0.27  qwen/qwen3-coder-30b-a3b-instruct
    0.09     0.35  deepseek/deepseek-v3.2-lite
    0.11     0.40  google/gemma-4-31b-it
```

###

Read the top line properly, because the number is easy to skim past. **One cent per million input tokens.** A million tokens is roughly three-quarters of a million words --- about ten novels. For a penny.

Now look at *which* models these are. Llama, Gemma, Qwen, Mistral, Nemotron, DeepSeek, `gpt-oss`. Every one of them is open weights. That is the point the last section made, showing up as a sorted list: the cheap end of this market is entirely made of models that more than one company is allowed to sell.

Notice too that output costs roughly three or four times input, everywhere, at every price level. Models read cheaply and write expensively. That is worth knowing before you ask for a long answer you were not going to read.

### Exercise 2

Give the cheap model the same job you gave Opus. In the bash Terminal:

<pre><code>curl -s https://openrouter.ai/api/v1/chat/completions \
  -H "Authorization: Bearer $OPENROUTER_API_KEY" \
  -H "Content-Type: application/json" \
  -d '{"model": "qwen/qwen3-coder-30b-a3b-instruct", "max_tokens": 400, "messages": [{"role": "user", "content": "Write R code to count rows per cut in the diamonds data set."}]}' \
  | python -c "import sys, json; d = json.load(sys.stdin); print(d['choices'][0]['message']['content']); print('--- cost:', d['usage']['cost'])"
</code></pre>

CP/CR.

```{r cheap-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 14)
```

###

```
library(ggplot2)
library(dplyr)

diamonds %>%
  count(cut)

--- cost: 0.0000237
```

###

The same answer as Opus, for **one two-hundred-and-fortieth of the price**.

Say that back plainly, because it is the most useful sentence in this tutorial: on this task, the model costing a fraction of a cent and the model costing half a cent produced work you cannot tell apart. Paying more bought you nothing at all.

That is not a trick or a fluke. Most of what anyone asks a model to do all day is this easy --- rename a variable, write the boilerplate, remind me of the argument order, turn this into a function. For that work the cheap models have been good enough for a while now, and the price difference between "good enough" and "the best there is" is a hundredfold.

### Exercise 3

You have now had three models write essentially the same code, and not one of them ran it. Do that part yourself.

In the R Terminal:

<pre><code>library(ggplot2)
library(dplyr)

diamonds |> count(cut)
</code></pre>

CP/CR.

```{r cheap-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

```
# A tibble: 5 × 2
  cut           n
  <ord>     <int>
1 Fair       1610
2 Good       4906
3 Very Good 12082
4 Premium   13791
5 Ideal     21551
```

###

There are the real counts, and they took less than a second to get.

Compare them once more against the free model's invented table from the last section, which claimed 68,146 Premium diamonds in a data set with 53,940 rows altogether. The cheap paid model did not make that mistake. Neither did Opus. But the reason you *know* none of them did is that you just ran the code.

**The habit is the deliverable here.** A model writes code; you run it. No price point removes that step, and no model is expensive enough to be trusted with it.

### Exercise 4

One last question, and it is the one that decides how you spend money for the rest of your career.

If the cheap model matched Opus on that task, why would anyone pay two hundred times more? Write down what you think, in a sentence or two. There is no wrong answer here --- we are after your guess before we give you ours.

```{r cheap-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Ours:

```
Because the task was easy. The gap probably shows up on hard problems,
long files, or when you need it to follow instructions exactly.
```

###

That is essentially right, and it is worth being precise about where the money goes. Roughly, as tasks get harder:

- **Free models** (`:free`) handle boilerplate and short scripts. They are rate limited, run by whichever company is giving them away today, and paid for with your prompts. Fine for practice, wrong for anything private.
- **Cheap open weights**, a few cents per million tokens. The workhorse tier, and where most of your work should live. Correct code for ordinary tasks, at a price you can stop thinking about.
- **Mid-priced models**, a few tens of cents. Better when the context is long, when a change spans several files, or when you need the output in an exact shape every single time.
- **Frontier models**, dollars per million tokens. Worth it for the problem you are genuinely stuck on, for reasoning that has to hold together over many steps, and for anything where being confidently wrong is expensive.

The gap between the ends of that list is real. It is just nothing like proportional to the price. You are not getting a hundred times more; you are getting somewhat better answers on the hardest tenth of your work, and paying for exclusivity on the rest.

###

So do not choose a model by reputation, and do not choose one from a benchmark table. Both of those are answering a different question than yours.

Choose it the way you just did. Take a task you actually have, run it on something cheap, and look at the result. Move up only when you can point at what the cheap model got wrong. That test costs a fraction of a cent and takes a minute, and it is the only evidence about *your* work that exists.

## Models From R
###

Everything so far has been about writing code. But the reason a data scientist cares about cheap models is bigger than that: **a model is something you can run on every row of a data set.**

### Exercise 1

In the R Terminal, run `library(ellmer)`. CP/CR.

```{r ellmer-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

<!-- ellmer must be in the PPBDS/devcontainers image before this ships; until then
     students hit "there is no package called 'ellmer'". Verified against ellmer
     0.4.2: chat_openrouter(model = ...) constructs, reads OPENROUTER_API_KEY,
     returns an R6 Chat object. -->

```
> library(ellmer)
```

###

**[ellmer](https://ellmer.tidyverse.org/)** is the tidyverse package for talking to AI models from R. It does the same job as the `curl` commands above, but hides the web request. You get one function per service, a chat object you can keep talking to, and it uses the same `OPENROUTER_API_KEY` you have been using all along.

It is not the only choice. **[httr2](https://httr2.r-lib.org/)** would let you build the request by hand, and Python has the `openai` library. But `ellmer` fits the way you already work in R.

### Exercise 2

Still in the R Terminal, create a chat and ask it something:

<pre><code>chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat("In one sentence, what is a tibble?")
</code></pre>

CP/CR.

```{r ellmer-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

```
> chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
> chat$chat("In one sentence, what is a tibble?")
User asks: "In one sentence, what is a tibble?" Provide one sentence
definition. Something like "In R, a tibble is a modern re-implementation of
data frames that is part of the tidyverse." Should answer in one sentence.
Alright.A **tibble** is a tidy-verse-style, modern reimplementation of R's
data frame that stores tabular data as a list of consistent-typed columns
with a clean, concise printing, and features straightforward subsetting.
```

###

`chat_openrouter()` found your key on its own. It read `OPENROUTER_API_KEY`, the same variable `aider` and `curl` used. One key, three tools.

Now look at what came back. The first few lines are not the answer. They are the model's *reasoning* --- the same `reasoning` field you saw in the JSON earlier, which `ellmer` prints too. You can see exactly where thinking stops and answering starts: `Alright.A **tibble** is`, with no space, because two separate pieces got stuck together.

Your output will not match ours word for word, and may be messier than ours. Free models are run by whichever company is giving them away that day, and those are not always the ones being careful.

Always name the model. `ellmer` has a default, and its own documentation warns that the default changes. Code that relies on it may start costing money one day without you touching it.

### Exercise 3

Now the thing you could not do by hand. In the R Terminal:

<pre><code>reviews <- c("the room was filthy", "staff went out of their way for us",
             "fine, but overpriced", "would absolutely stay again")

chat <- chat_openrouter(model = "openai/gpt-oss-20b:free")
chat$chat(paste("Label each line positive, negative, or mixed. One label per line, nothing else:",
                paste(reviews, collapse = "\n")))
</code></pre>

CP/CR.

```{r ellmer-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

Ours, with the reasoning trimmed --- the last four lines are what it offered as the answer:

```
1. "the room was filthy" -> Negative (clear negative).
2. staff ... it's a compliment about staff, positive.
3. "fine, but overpriced" -> Mixed?  ...
We need just one per line, no additional text.
negative
positive
mixed
 perhatian positif
```

###

Three labels are right. The fourth is `perhatian positif` --- Indonesian for "positive attention." We asked for one label per line and nothing else. We got a paragraph of thinking, three correct labels, and then a word in the wrong language.

This one result shows both why cheap models matter and why you cannot trust them blindly.

Four hotel reviews is a toy. Four *hundred thousand* is a real data set, and until recently the only way to turn that much writing into a usable column was to pay people to read it. Four hundred thousand short reviews is roughly 20 million tokens: about **$1.40** on the Qwen model, about **$100** on Opus.

But a column containing `perhatian positif` is not data, it is a bug. Two habits fix this, and professionals use both:

1. **Ask for a fixed shape.** `ellmer` can require the model to return one of exactly three labels instead of free writing. A wrong answer then becomes an error you can catch.
2. **Check anyway.** Count that you got back as many labels as you sent rows, and that every label is one you expected.

Neither is optional once the data is big. With four reviews you can spot the bad one. With four hundred thousand you cannot.

You have credits now, so you can test the other half of that claim rather than take our word for it. Run the same thing again with `model = "qwen/qwen3-coder-30b-a3b-instruct"`. On a task this small it will cost you well under a hundredth of a cent, and `perhatian positif` will almost certainly not come back. Cheap is not the same as free, and this is one of the places you can see the difference.

## What It Cost
###

You spent money today for the first time in this course. This section is the bill.

### Exercise 1

In the bash Terminal, check what you have spent:

<pre><code>curl -s https://openrouter.ai/api/v1/key -H "Authorization: Bearer $OPENROUTER_API_KEY" | python -c "import sys, json; d = json.load(sys.stdin)['data']; print('spent so far:', d['usage']); print('free tier:', d['is_free_tier'])"
</code></pre>

CP/CR.

```{r paying-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
spent so far: 0.0061
free tier: False
```

###

Six tenths of a cent. Yours will differ a little, but not by much.

That is the entire cost of running a frontier model, a cheap model, and a pile of free ones, plus everything you did from R. Of the $9.45 the purchase actually bought you, you have used well under one percent. At this rate the balance outlives the course by a wide margin.

Notice the habit in that command, too: the reply had fourteen fields and we asked for two. An API hands you everything it has. A good program takes only what it came for.

### Exercise 2

Protect the balance before you leave. Go back to <https://openrouter.ai/workspaces/default/keys>, edit your `codespace` key, and set a **credit limit** of `$2`.

Copy and paste the name of the key and the limit you set.

```{r paying-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

<!-- VERIFY: that an existing key can have a credit limit added after creation on
     the current OpenRouter keys page. If the limit can only be set at creation
     time, this exercise becomes "delete the key and make a new one with a
     limit," and Exercise 3's key rotation note must move up. -->

```
codespace
$2
```

###

Once a key has real money behind it, a key that leaks costs you real money. A limit turns the worst case from "my balance" into "two dollars."

It also protects you from yourself, which is the more likely accident. You now know that everything sent to a model is re-sent with every message; put that inside a loop over a data frame with four hundred thousand rows and a mistake gets expensive quickly. Professionals cap keys as a matter of course, and they cap the ones agents use hardest.

The habit generalises well past OpenRouter. Any credential that can spend money should have the smallest limit that still lets it do its job.

### Exercise 3

Commit and push your work.

In the bash Terminal, run `git log --oneline`. CP/CR.

```{r paying-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
models-and-money $ git log --oneline
7d21ab8 (HEAD -> main, origin/main) Add model comparison work
```

###

Nothing you committed contains your key. That is worth checking rather than assuming. The key lived in an environment variable and in GitHub's encrypted store, and never in a file. Copy that arrangement on every project you ever work on.

## Summary
###

You took the assistant away and found one ordinary web request underneath: a model name, a list of messages, and a key. Everything built on top of an AI model is built on that.

###

You saw prices as data rather than marketing, and found a ninety-three-fold gap between a cheap model and an expensive one. The reason is competition, not quality. Anyone can run a published model, so its price falls to the cost of running it. Only one company can run a closed model, so it charges what it likes.

That is why the sensible habit is a cheap model for most work and an expensive one for the problem you are truly stuck on.

###

Then you tested it rather than believing it. The same prompt went to a frontier model and to a model costing a fraction of a cent, and the answers were indistinguishable --- one at half a cent, one at two hundred-and-fortieth of that.

What the expensive model was better at turned out to be something else entirely. Given a question it could not answer, it declined instead of inventing a table of numbers the way the free model did. On work where being confidently wrong is expensive, that is most of what the money buys.

So pick a model the way you picked one today: run your own task on something cheap, and move up only when you can point at what it got wrong.

###

You called a model from R with **ellmer** and used it for the job nothing else can do: turning ordinary writing into a column of data. You also saw it get one of four labels wrong, which is why checking the output is part of the job.

###

And you spent, in total, well under a cent.

The credit card was a real requirement and this was the only place in the course that asks for it. What it bought was not the tutorial --- it was the ability to check our claims instead of taking them. The remaining balance is yours, it is capped at $2 per key, and it will outlast this course.

You now know what a model costs, what it costs to be careless with one, and how to find out whether the expensive one is worth it for the work you actually do. That last one is the skill. The model names in this tutorial will all be gone in a year.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
