---
title: "Introduction"
author: David Kane
tutorial:
  id: "00-introduction"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "The first tutorial in the sequence. Introduces how the tutorials work and gets your working environment set up."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

This is the first tutorial in the sequence. It introduces how the tutorials work and gets your working environment set up.

### Exercise 1

Create and connect to a repo called `00-introduction`, then open it as your workspace.

Copy and paste the URL for your `00-introduction` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like:

````
https://github.com/ppbds-student/00-introduction
````

###

TODO: knowledge drop.

## Summary
###

This tutorial introduced how the tutorials work and got your working environment set up.

### Exercise 1

Commit and push any remaining changes. Copy and paste the URL for your `00-introduction` repository.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look something like:

````
https://github.com/ppbds-student/00-introduction
````

###

TODO: knowledge drop.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
