---
title: "GitHub Advanced"
author: David Kane
tutorial:
  id: "09-github-2"
output:
  learnr::tutorial:
    progressive: yes
    'allow_skip:': yes
runtime: shiny_prerendered
description: Independence from connect-repo. Create a repo by hand with gh, push with -u, clone it onto a fresh machine, pull changes back, and delete a Codespace while your work survives on GitHub.
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In the GitHub Introduction tutorial, you opened up the `connect-repo` script and ran its key commands once, against a throwaway repo. This tutorial is the independence drill. `connect-repo` is a convenience script that lives in `codespace-starter` --- a repo you do not control, and a script that might change or disappear. By the end of this tutorial, that will not matter: you will have run the entire repo lifecycle --- create, connect, work, push, clone onto a fresh machine, pull, and dispose --- with your own hands.

So, unlike every other tutorial: **do not run `connect-repo` today.** That is the point.

Some material is from [*R for Data Science (2e)*](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund. The most useful reference for Git/GitHub is [*Happy Git and GitHub for the useR*](https://happygitwithr.com/).

## A repo by hand
###

Everything in this section happens in your current Codespace. You will make the repo this tutorial is named for --- `github-advanced` --- using only `gh` and `git`.

### Exercise 1

Open a new bash Terminal and run `cd /workspaces`. (New terminals may open inside the repo from an earlier tutorial --- that auto-jump is a `connect-repo` convenience; from now on, we `cd` ourselves.) Then run `gh auth status`.

If it reports that you are **not** logged in, run these two commands --- they are steps 1 and 2 from `connect-repo --help` --- and then run `gh auth status` again:

````
unset GITHUB_TOKEN GH_TOKEN
gh auth login --hostname github.com --git-protocol https --web
````

CP/CR.

```{r a-repo-by-hand-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

````
workspaces $ gh auth status
github.com
  ✓ Logged in to github.com account your-username
  - Active account: true
  - Git operations protocol: https
  - Token: gho_************************************
  - Token scopes: 'gist', 'read:org', 'repo', 'workflow'
````

###

You were probably already signed in --- `connect-repo` did that once, in some earlier session, and the credentials stick around. What matters is that you now know how the sign-in happens without the script. Every `git` and `gh` command you run --- and every AI agent that pushes code on your behalf --- uses exactly these credentials.

### Exercise 2

Create this tutorial's repo by hand. From `/workspaces`, run:

````
gh repo create github-advanced --public --clone
````

Then run `ls`. CP/CR.

```{r a-repo-by-hand-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

````
workspaces $ gh repo create github-advanced --public --clone
✓ Created repository your-username/github-advanced on github.com
  https://github.com/your-username/github-advanced
Cloning into 'github-advanced'...
warning: You appear to have cloned an empty repository.
workspaces $ ls
codespace-starter  github-advanced
workspaces $
````

Your `ls` may also show repos from earlier tutorials. That is fine.

###

Why `--public`? GitHub repos are either **public** or **private**. A public repo is visible to anyone on the internet --- not just people you share the link with. A private repo is visible only to you and collaborators you explicitly invite. We use public repos in this course, which is also what lets GitHub Pages host your work for free. For work that should stay private, `gh repo create <name> --private --clone` is the only change you need.

### Exercise 3

Run `cd github-advanced` and then `git remote -v`. CP/CR.

```{r a-repo-by-hand-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

````
workspaces $ cd github-advanced
github-advanced $ git remote -v
origin  https://github.com/your-username/github-advanced (fetch)
origin  https://github.com/your-username/github-advanced (push)
github-advanced $
````

###

Same wiring you inspected in the GitHub Introduction tutorial: `origin` is the nickname for the GitHub copy, and the clone set it up for you. Create, clone, wire --- three jobs, one command, no script.

### Exercise 4

Point VS Code at your new repo: `File -> Open Folder...`, then open `/workspaces/github-advanced`. The window will reload and stop this tutorial. That is OK! Restart it the same way you started it: click **R Tutorials** in the Activity Bar, find this tutorial under **vscode.tutorials**, and click the arrow to launch it. All of your saved answers will still be there.

Then, in a new bash Terminal, run these two commands, one at a time:

````
cp /workspaces/codespace-starter/.devcontainer/AGENTS.md .
ls
````

CP/CR.

```{r a-repo-by-hand-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

````
github-advanced $ cp /workspaces/codespace-starter/.devcontainer/AGENTS.md .
github-advanced $ ls
AGENTS.md
github-advanced $
````

###

Count what you have now reproduced by hand: the sign-in, the repo creation, the clone, the window switch, and `AGENTS.md` --- which turns out to be just a file that `connect-repo` copies, telling AI assistants how to work with you. The script's remaining conveniences are cosmetic. You no longer need it.

### Exercise 5

Give the repo something worth keeping, set up the way you learned in the Quarto tutorial: create `analysis.qmd` with the title "GitHub Advanced", your name as author, `execute: echo: false` in the YAML, a setup chunk loading **tidyverse** with `#| message: false`, and a second chunk holding an AI-generated plot using a built-in data set. Also create a `.gitignore` containing `*_files`. Render with `Cmd/Ctrl + Shift + K`.

In the R Terminal, run `tutorial.helpers::show_file("analysis.qmd", chunk = "Last")`. CP/CR.

```{r a-repo-by-hand-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

Your plot code will differ from ours. Something like:

````
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7) +
  labs(title = "Engine Displacement vs. Highway MPG",
       x = "Engine Displacement (L)",
       y = "Highway MPG",
       color = "Car Class")
````

###

Nothing in this exercise was new, and it probably took you only a few minutes --- the YAML header, setup chunk, `.gitignore`, and an AI plot are muscle memory by now. That speed is the payoff of repetition.

### Exercise 6

Before committing, see what Git sees. Run `git status`. CP/CR.

```{r a-repo-by-hand-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

````
github-advanced $ git status
On branch main

No commits yet

Untracked files:
  (use "git add <file>..." to include in what will be committed)
        .gitignore
        AGENTS.md
        analysis.html
        analysis.qmd

nothing added to commit but untracked files present (use "git add" to track)
````

###

Note what is *not* listed: `analysis_files`, because your `.gitignore` is already working. `git status` shows exactly the same information as the Source Control pane --- untracked files there wear a "U" --- just in text form. The pane is a visual wrapper around this command. Knowing the text version means you know what the buttons actually do, and it is how you check up on an AI agent's work in your repo.

### Exercise 7

Make the first push. Run these three commands, one at a time:

````
git add -A
git commit -m "First analysis"
git push -u origin main
````

CP/CR.

```{r a-repo-by-hand-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

````
github-advanced $ git add -A
github-advanced $ git commit -m "First analysis"
[main (root-commit) 1a2b3c4] First analysis
 4 files changed, 152 insertions(+)
 create mode 100644 .gitignore
 create mode 100644 AGENTS.md
 create mode 100644 analysis.html
 create mode 100644 analysis.qmd
github-advanced $ git push -u origin main
Enumerating objects: 6, done.
Writing objects: 100% (6/6), 24.31 KiB | 8.10 MiB/s, done.
To https://github.com/your-username/github-advanced
 * [new branch]      main -> main
branch 'main' set up to track 'origin/main'.
github-advanced $
````

###

The new piece is `-u` (for `--set-upstream`). A brand-new repo's `main` branch does not yet know which remote branch it corresponds to, so the very first push must say so: `-u origin main` means "push to `origin`, and remember this pairing." After that, plain `git push` works forever. (Some setups configure Git to do this automatically, in which case plain `git push` succeeds even the first time --- `-u` is still the portable habit that works everywhere.) Note also `root-commit` in the commit output: this is the first commit in the repo's entire history.

### Exercise 8

In your browser, go to your `github-advanced` repository page and confirm all four files are there. Copy and paste the URL.

```{r a-repo-by-hand-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

````
https://github.com/your-username/github-advanced
````

###

Notice that next to each file, GitHub shows the commit message from the last time that file changed. Good commit messages leave a breadcrumb trail visible right on the repo's main page, not just buried in `git log`.

## A second desk
###

The real test of independence is a machine you have never touched. On your own laptop, in a colleague's Codespace, on a fresh cloud server --- there is no `connect-repo` there, and there are no credentials. Codespaces make fresh machines cheap, so let's take the test.

### Exercise 1

In a **new browser tab**, go to [github.com/codespaces](https://github.com/codespaces). Click the green **New codespace** button, choose the repository `PPBDS/codespace-starter`, leave the branch as `main`, and click **Create codespace**.

Keep the tutorial tab open! You now have two Codespaces: the one running this tutorial, and a fresh machine. The first boot takes a few minutes.

Copy and paste the new Codespace's URL (from its browser tab, ending in `.github.dev`).

```{r a-second-desk-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

````
https://zany-computing-machine-vpvrv9p59jxpc66v6.github.dev/
````

###

While you waited, GitHub built a computer from scratch, following the instructions in `.devcontainer/devcontainer.json` inside `codespace-starter`: which R version to install, which packages to pre-load, which editor extensions to add. That is why every Codespace arrives with R, tidyverse, and Quarto already configured --- none of it happens by accident. The first startup is the slowest because all of that setup work is happening; reopening an existing Codespace is much faster.

### Exercise 2

In the **new** Codespace, open a fresh bash Terminal and run the sign-in ritual by hand --- this time it is genuinely needed, because this machine has never met you. Run these commands, one at a time:

````
unset GITHUB_TOKEN GH_TOKEN
gh auth login --hostname github.com --git-protocol https --web
gh auth status
````

For the login, follow the prompts: press `Enter` to accept the browser flow, copy the one-time code, click Authorize on the GitHub page, and return. CP/CR the final `gh auth status` output.

```{r a-second-desk-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

````
codespace-starter $ gh auth status
github.com
  ✓ Logged in to github.com account your-username
  - Active account: true
  - Git operations protocol: https
  - Token: gho_************************************
  - Token scopes: 'gist', 'read:org', 'repo', 'workflow'
````

###

Why the `unset`? A Codespace starts with a built-in token that acts as the *Codespace*, not as *you* --- it can read `codespace-starter` and nothing more. Clearing it forces `gh` to use your real login instead. This whole ritual happens once per machine. On your own laptop, the equivalent is installing `gh` and running `gh auth login` --- same command, no `unset` needed.

### Exercise 3

Still in the new Codespace, connect to your existing repo. Run these commands, one at a time:

````
cd /workspaces
gh repo clone github-advanced
cd github-advanced
ls
````

CP/CR.

```{r a-second-desk-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

````
codespace-starter $ cd /workspaces
workspaces $ gh repo clone github-advanced
Cloning into 'github-advanced'...
remote: Enumerating objects: 6, done.
remote: Counting objects: 100% (6/6), done.
Receiving objects: 100% (6/6), 24.31 KiB | 3.04 MiB/s, done.
workspaces $ cd github-advanced
github-advanced $ ls
AGENTS.md  analysis.html  analysis.qmd
github-advanced $
````

###

This is the connect-to-an-existing-repo case: no creation, just a clone. `gh repo clone github-advanced` is shorthand for `git clone https://github.com/your-username/github-advanced.git` --- the plain-`git` form works on any machine, even one without `gh`. And notice what the clone does *not* contain: `analysis_files`. It was ignored, so it was never pushed --- your `.gitignore` decision from the first desk travels with the repo.

### Exercise 4

Make a change on this second desk. Open `analysis.qmd` (use the Explorer, or `File -> Open Folder...` to `/workspaces/github-advanced` first if you prefer a full window). Below the last chunk, add the sentence:

````
This line was written on a second machine.
````

Save. Then run these three commands, one at a time:

````
git add -A
git commit -m "Add a line from the second machine"
git push
````

CP/CR.

```{r a-second-desk-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

````
github-advanced $ git add -A
github-advanced $ git commit -m "Add a line from the second machine"
[main 5d6e7f8] Add a line from the second machine
 1 file changed, 2 insertions(+)
github-advanced $ git push
Enumerating objects: 5, done.
Writing objects: 100% (3/3), 341 bytes | 341.00 KiB/s, done.
To https://github.com/your-username/github-advanced
   1a2b3c4..5d6e7f8  main -> main
github-advanced $
````

###

Plain `git push` worked --- no `-u` needed. When you clone a repo that already has commits, Git wires the upstream tracking automatically. The `-u` flag is only for a branch's very first push, as in the last section.

### Exercise 5

Return to your **original** Codespace --- the tab running this tutorial. Its copy of `github-advanced` does not know about the new commit yet. In a bash Terminal there (make sure `pwd` says `/workspaces/github-advanced`), run `git status` and then `git pull`, one at a time. CP/CR.

```{r a-second-desk-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

````
github-advanced $ git status
On branch main
Your branch is up to date with 'origin/main'.

nothing to commit, working tree clean
github-advanced $ git pull
remote: Enumerating objects: 5, done.
Unpacking objects: 100% (3/3), 321 bytes | 321.00 KiB/s, done.
From https://github.com/your-username/github-advanced
   1a2b3c4..5d6e7f8  main       -> origin/main
Updating 1a2b3c4..5d6e7f8
Fast-forward
 analysis.qmd | 2 ++
 1 file changed, 2 insertions(+)
github-advanced $
````

###

Look closely: `git status` claimed you were "up to date with 'origin/main'" --- and it was wrong. `git status` never contacts GitHub; it only compares against the last snapshot of `origin` it has seen. `git pull` actually asks GitHub, fetches the new commit, and merges it in. This is the other half of syncing: **push sends, pull receives**. The Source Control pane's "Sync" button does both at once. Whenever more than one desk touches a repo --- a second machine, a collaborator, an AI agent working in the cloud --- pull before you work.

## Disposable desks
###

Two desks, one filing cabinet. Time to prove which one matters.

### Exercise 1

Go to [github.com/codespaces](https://github.com/codespaces). You should see (at least) two active Codespaces. Copy and paste the text of the table.

```{r disposable-desks-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

Your submission should show two entries, something like:

````
ppbds-student
PPBDS/codespace-starter
automatic funicular
main
Active

ppbds-student
PPBDS/codespace-starter
zany computing machine
main
Active
````

###

Every active Codespace counts against your free Codespace hours. Get in the habit of visiting this page and cleaning up --- stopped Codespaces cost storage, deleted ones cost nothing.

### Exercise 2

Delete the second Codespace --- the fresh machine from the last section. Click the `...` next to it and select **Delete**. (Be sure you are deleting the right one: not the Codespace running this tutorial.)

Then, back in your original Codespace, run `git log --oneline` in the `github-advanced` Terminal. CP/CR.

```{r disposable-desks-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

````
github-advanced $ git log --oneline
5d6e7f8 (HEAD -> main, origin/main) Add a line from the second machine
1a2b3c4 First analysis
github-advanced $
````

###

The machine that wrote the second commit no longer exists --- and the commit does not care. It lives in the filing cabinet on GitHub and on every desk that has pulled it. This is why we commit and push so often: not to save work *to* a Codespace, but to save it *through* the Codespace to GitHub. Desks are disposable. The cabinet is permanent.

## Summary
###

You ran the entire repo lifecycle without `connect-repo`: signed in with `gh auth login`, created a repo with `gh repo create --public --clone`, inspected its wiring with `git remote -v`, pushed a first commit with `git push -u origin main`, cloned onto a brand-new machine with `gh repo clone`, pushed from there with plain `git push`, brought the change home with `git pull`, and deleted a Codespace without losing a byte.

These are the commands you now own:

````
gh auth login        # once per machine
gh repo create <name> --public --clone
gh repo clone <name>
git status
git add -A
git commit -m "message"
git push -u origin main   # first push of a new branch
git push                  # every push after that
git pull
git remote -v
git log --oneline
````

`connect-repo` is a convenience, not a dependency. If it disappeared tomorrow, you would be fine.

### Exercise 1

Copy and paste the URL for your `github-advanced` repository. This is the last question in the tutorial.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

````
https://github.com/your-username/github-advanced
````

###

The most useful reference for Git/GitHub is [*Happy Git and GitHub for the useR*](https://happygitwithr.com/). Refer to that book --- or ask AI, whose git knowledge is excellent --- whenever you have a problem.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
