---
title: "Our Codespace Starter"
author: David Kane
description: "A tour of the devcontainer.json behind codespace-starter itself --- the custom PPBDS image, the VS Code extensions array, the settings that control auto-save, plots, and Git commits, and the retired postCreateCommand whose recipe is kept dormant in a comment."
format: live-html
engine: knitr
toc: true
learnr2:
  ordering: 14
---

{{< include _extensions/r-wasm/live/_knitr.qmd >}}

## Information
###

```{r}
#| label: information-1
#| echo: false
learnr2::student_info()
```

## Introduction
###

In the Devcontainers tutorial you examined a generic Python devcontainer and wrote a minimal R devcontainer from scratch. Now you will look at the devcontainer behind the Codespace you are working in right now. `codespace-starter` uses a custom image built specifically for this course, plus a detailed `settings` block that configures all of it.

### Exercise 1

You should be doing this tutorial in a repo named `our-codespace-starter`. If you are not, create one and then connect to it. You may need to restart this tutorial after you do so.

Copy and paste the URL for your `our-codespace-starter` repository.

```{r}
#| label: introduction-1
#| echo: false
learnr2::question(
  "Copy and paste the URL for your our-codespace-starter repository.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
https://github.com/ppbds-student/our-codespace-starter
```

###

This new repo is empty except for a `.git` folder --- the `devcontainer.json` that actually configured this codespace lives somewhere else entirely, which you'll track down in a couple of exercises.

### Exercise 2

<!-- DK: Delete the next few questions? They are the same as in 10. Bug or feature? -->

Copy and paste the URL for your codespace.

```{r}
#| label: introduction-2
#| echo: false
learnr2::question(
  "Copy and paste the URL for your codespace.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
https://congenial-space-broccoli-7qpj5xvw9xhpqj9.github.dev/
```

###

Every codespace runs on its own remote machine that GitHub provisions for you --- this URL is that machine's unique address, and it stays the same for the rest of this tutorial.

### Exercise 3

In the bash Terminal, run `cd ../codespace-starter`, then `ls .devcontainer`. CP/CR.

```{r}
#| label: introduction-3
#| echo: false
learnr2::question(
  "In the bash Terminal, run cd ../codespace-starter, then ls .devcontainer. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
our-codespace-starter $ cd ../codespace-starter
codespace-starter $ ls .devcontainer
AGENTS.md  STUDENT_WORKFLOW.md  connect-repo.sh  devcontainer.json  setup.sh  tests  welcome.sh
```

###

Your new repo and `codespace-starter` sit side by side under `/workspaces` in this same Codespace. `codespace-starter` is the checkout GitHub actually read to build this machine, including the `devcontainer.json` you'll spend the rest of this tutorial reading.

## Codespace Starter Devcontainer
###

This section walks through `codespace-starter’`s `.devcontainer/devcontainer.json` field by field, examining the custom image, VS Code extensions, settings, the `onCreateCommand` behind the ready banner, and the retired `postCreateCommand`. Unlike the minimal configurations from the Devcontainers tutorial, this setup customizes almost everything to guarantee an identical experience for every student. By the end, you'll understand each top-level field and manually install a package to see what installing by hand looks like now that everything the course needs ships baked into the image.

### Exercise 1
 
In the bash Terminal, run `cd /workspaces/codespace-starter`. In the Explorer, open `.devcontainer/devcontainer.json` to see the whole file on your screen. Find the line beginning with `"name"`. Copy and paste it.

```{r}
#| label: codespace-starter-1
#| echo: false
learnr2::question(
  "In the bash Terminal, run cd /workspaces/codespace-starter. In the Explorer, open .devcontainer/devcontainer.json to see the whole file on your screen. Find the line beginning with \"name\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"name": "Preceptor's Primer",
```

###

The `name` field is purely cosmetic --- it is the label GitHub shows for this configuration in the Codespace-creation picker and the VS Code window title.

### Exercise 2

Find the line beginning with `"image"`. Copy and paste it.

```{r}
#| label: codespace-starter-2
#| echo: false
learnr2::question(
  "Find the line beginning with \"image\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"image": "ghcr.io/ppbds/devcontainer:1.1.5",
```

###

Your `codespace-starter` uses a pinned version of a custom image, `ghcr.io/ppbds/devcontainer`, so every student gets the exact same setup, pre-loaded with Quarto, R, Python, the four course tutorial packages, and various AI tools. We will explore images --- how they work and how to build them --- in a future tutorial.

### Exercise 3

Find the `"extensions"` array. Find the line containing `"reditorsupport.r"`. Copy and paste it.

```{r}
#| label: codespace-starter-3
#| echo: false
learnr2::question(
  "Find the \"extensions\" array. Find the line containing \"reditorsupport.r\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"reditorsupport.r@3.0.1",
```

###

`reditorsupport.r` is the extension that gives VS Code R syntax highlighting, code completion, and the R Terminal integration you've been using all course. The `@3.0.1` **pins** it to one exact version. Without a pin, the Marketplace would silently update every student's Codespace the day a new release appeared --- and the comment directly above this line explains why that matters here: version 3.0 of this extension changed how it attaches to a running R session (which is what gets plots into the plot pane), and it needs a matching R package, `sess`, that the course's image bakes in alongside this exact version. An uninvited update would have broken plotting mid-semester. Pinning is how the course decides *when* to take a change like that, rather than letting the Marketplace decide for it.

### Exercise 4

Find the line containing `"quarto.quarto"`. Copy and paste it.

```{r}
#| label: codespace-starter-4
#| echo: false
learnr2::question(
  "Find the line containing \"quarto.quarto\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"quarto.quarto",
```

###

`quarto.quarto` is what lets VS Code render and preview Quarto documents directly in the editor.

### Exercise 5

One extension is conspicuously missing from the `"extensions"` array: `PPBDS.vscode-r-tutorials`, the panel this very tutorial is running in. Find the `NOTE` comment just above the array that explains why it is not listed. Copy and paste the comment's first two lines.

```{r}
#| label: codespace-starter-5
#| echo: false
learnr2::question(
  "One extension is conspicuously missing from the \"extensions\" array: PPBDS.vscode-r-tutorials, the panel this very tutorial is running in. Find the NOTE comment just above the array that explains why it is not listed. Copy and paste the comment's first two lines.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
// NOTE — PPBDS.vscode-r-tutorials is deliberately NOT in this list.
// We publish it to Open VSX, and this list can only install from the
```

###

`PPBDS.vscode-r-tutorials` is the one extension unique to this course --- and it is deliberately absent from this list. The `"extensions"` array can only install from the Microsoft Marketplace, but we publish the extension to Open VSX. So it is baked into the image instead (the `RT_EXT_VERSION` build argument in the PPBDS/devcontainers Dockerfile), pre-extracted so its Activity Bar icon exists from the very first window paint. You are using it right now all the same.

### Exercise 6

Just below `"mechatroner.rainbow-csv"`, find the comment that begins `ALSO DELIBERATELY ABSENT`. Copy and paste its first five lines.

```{r}
#| label: codespace-starter-6
#| echo: false
learnr2::question(
  "Just below \"mechatroner.rainbow-csv\", find the comment that begins ALSO DELIBERATELY ABSENT. Copy and paste its first five lines.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
// ALSO DELIBERATELY ABSENT (dropped 2026-08-20, David): the Python
// extensions (ms-python.python, ms-toolsai.jupyter). Students use
// only a handful of Activity Bar icons, and these two added Python/
// Jupyter icons plus Welcome-page walkthrough cards for tooling the
// course doesn't teach interactively. Removing them does NOT break
```

###

Standard JSON doesn't allow comments, but `devcontainer.json` is actually **JSONC** (JSON with Comments) --- that's what lets a `//` note like this one record a decision right next to the code it affects. This one explains an *absence*: the Python and Jupyter extensions used to be in the list, and were removed to keep the Activity Bar down to the few icons the course actually uses. Read the rest of the comment and notice what it promises: Python still works. `quarto render` runs the Python chunks in a `.qmd` through the image's Jupyter engine from the command line, with no editor extension involved. What went away is only the in-editor layer --- notebook tabs, run buttons, and code completion for Python.

### Exercise 7

Find the lines containing `"ritwickdey.LiveServer"`, `"tomoki1207.pdf"`, and `"mechatroner.rainbow-csv"`. Copy and paste all three.

```{r}
#| label: codespace-starter-7
#| echo: false
learnr2::question(
  "Find the lines containing \"ritwickdey.LiveServer\", \"tomoki1207.pdf\", and \"mechatroner.rainbow-csv\". Copy and paste all three.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"ritwickdey.LiveServer",
"tomoki1207.pdf",
"mechatroner.rainbow-csv"
```

###

Three quality-of-life extensions: `ritwickdey.LiveServer` previews HTML files, `tomoki1207.pdf` opens PDFs inside the editor, and `mechatroner.rainbow-csv` colors CSV columns to make them easier to read. Notice what's missing from this whole array: no AI-assistant extensions, because those tools run from the terminal instead, where they work on every device and render properly inside a browser-based Codespace.

### Exercise 8

Find the line containing `"r.consolePath"`. Copy and paste it.

```{r}
#| label: codespace-starter-8
#| echo: false
learnr2::question(
  "Find the line containing \"r.consolePath\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"r.consolePath": "/usr/local/bin/arf",
```

###

`arf` is a faster stand-in for the default R Terminal --- it starts quicker and handles pasted multi-line code more reliably. `/usr/local/bin/arf` is a fixed location the image always installs it at, so this setting keeps working even if the image changes internally.

### Exercise 9

Find the line containing `"r.plot.backend"`. Copy and paste it.

```{r}
#| label: codespace-starter-9
#| echo: false
learnr2::question(
  "Find the line containing \"r.plot.backend\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"r.plot.backend": "httpgd",
```

###

By default, R opens plots in a separate window that would be invisible in a Codespace. Naming `httpgd` here sends plots to the **httpgd** tool instead, which streams them into a panel inside VS Code. (The extension's own default, `auto`, would pick whichever plot device it happens to find; naming one keeps every student on the same device.)

### Exercise 10

Find the line containing `"r.source.echo"`. Copy and paste it.

```{r}
#| label: codespace-starter-10
#| echo: false
learnr2::question(
  "Find the line containing \"r.source.echo\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"r.source.echo": true,
```

###

The Run Source button (▷ in the editor title bar) normally runs a file silently, showing only the final output. This setting makes it run with `echo = TRUE` instead, so each line prints as it executes.

### Exercise 11

Find the line containing `"git.enableSmartCommit"`. Copy and paste it.

```{r}
#| label: codespace-starter-11
#| echo: false
learnr2::question(
  "Find the line containing \"git.enableSmartCommit\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"git.enableSmartCommit": true,
```

###

Smart commit automatically includes all changed files in the commit when you click the Commit button in the Source Control sidebar --- you do not need to run `git add` separately. This is why clicking Commit in the `codespace-starter` Codespace works without an extra confirmation step, while the Jupyter Notebook devcontainer setup required one.

### Exercise 12

Find the line containing `"files.autoSave"`. Copy and paste it.

```{r}
#| label: codespace-starter-12
#| echo: false
learnr2::question(
  "Find the line containing \"files.autoSave\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"files.autoSave": "afterDelay",
```

###

With `"afterDelay"`, VS Code automatically saves open files a short time after you stop typing. This prevents the most common beginner mistake in coding environments: running code that still reflects the last-saved version rather than the current edits.

### Exercise 13

Find the line containing `"editor.wordWrap"`. Copy and paste it.

```{r}
#| label: codespace-starter-13
#| echo: false
learnr2::question(
  "Find the line containing \"editor.wordWrap\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"editor.wordWrap": "on",
```

###

Word wrap causes long lines to wrap to the next line visually rather than extending off the right edge of the screen. Without it, long comment blocks or URLs in code appear truncated on smaller screens.

### Exercise 14

Find the line containing `"workbench.editor.enablePreview"`. Copy and paste it.

```{r}
#| label: codespace-starter-14
#| echo: false
learnr2::question(
  "Find the line containing \"workbench.editor.enablePreview\". Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"workbench.editor.enablePreview": false
```

###

By default, VS Code opens files in "preview" mode --- clicking a second file replaces the first tab, so beginners often think their file disappeared. Setting this to `false` means every file opens in a permanent tab.

### Exercise 15

Now leave the `"settings"` block and scroll back up. Find the `"onCreateCommand"` field. Copy and paste that one line.

```{r}
#| label: codespace-starter-15
#| echo: false
learnr2::question(
  "Now leave the \"settings\" block and scroll back up. Find the \"onCreateCommand\" field. Copy and paste that one line.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
"onCreateCommand": "bash ${containerWorkspaceFolder}/.devcontainer/setup.sh",
```

###

You have already met the result of this line --- it is behind the "✅ YOUR CODESPACE IS READY" banner that greeted you in your first terminal. An `onCreateCommand` runs once, when the container is first built; for this repo that happens during the prebuild, before you ever see the Codespace. The value is the command: `bash` running `setup.sh`, with `${containerWorkspaceFolder}` standing in for the absolute path of the checkout. Open `setup.sh` and you will find it does three small things: installs the `connect-repo` command, sets the short prompt you have been typing at all along, and adds a few lines to `~/.bashrc` that print the banner the first time a terminal opens. Nothing prints during the build itself --- its output goes to a log nobody reads --- which is why the banner is printed by your shell, later, rather than by this command directly. (Earlier versions of the starter used a `postAttachCommand` instead, which runs in a *visible* terminal every time the editor attaches. It worked, but it opened a second terminal next to the one Codespaces already gives you, and students kept asking why there were two.)

### Exercise 16

You might expect a `"postCreateCommand"` field --- a command that runs every time a Codespace is created. There isn't one. Find the comment that begins `NO postCreateCommand right now`. Copy and paste its first five lines.

```{r}
#| label: codespace-starter-16
#| echo: false
learnr2::question(
  "You might expect a \"postCreateCommand\" field -- a command that runs every time a Codespace is created. There isn't one. Find the comment that begins NO postCreateCommand right now. Copy and paste its first five lines.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
// NO postCreateCommand right now — RETIRED 2026-07-30. All course
// packages, including primer.tutorials, ship at whatever versions the
// image baked; the live at-create refresh of primer.tutorials was removed
// because the baked version is recent enough, making creates faster and
// GitHub-independent.
```

###

There *used* to be a `postCreateCommand` here that refreshed the course packages from GitHub every time a Codespace was created. It was retired: the four course packages now ship at whatever versions the image baked, and updates reach students through a new image release plus a pin bump --- the same `"image"` line you found in Exercise 2. That makes Codespace creation faster and independent of GitHub being reachable.

Read the rest of the comment, though. The recipe was kept on purpose, dormant, in case course content ever again needs to move faster than image releases --- along with hard-won notes on *why* it must be `postCreateCommand` rather than `onCreateCommand` if revived --- the opposite choice from the setup script above, and for a reason: setup can be frozen into the prebuild, a package refresh must not be. Deleting working knowledge is cheap; re-learning it is not. Either way, the packages only cover the course --- anything else you need, you install yourself, by hand or by asking AI, which is exactly what the next few exercises do.

### Exercise 17

In the R Terminal, run `install.packages("praise")`. CP/CR.

```{r}
#| label: codespace-starter-17
#| echo: false
learnr2::question(
  "In the R Terminal, run install.packages(\"praise\"). CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
> install.packages("praise")
Installing package into '/home/ppbds-student/R/library'
(as 'lib' is unspecified)
trying URL 'https://cloud.r-project.org/src/contrib/praise_1.0.0.tar.gz'
Content type 'application/x-gzip' length 6431 bytes
downloaded 6431 bytes

* installing *source* package 'praise' ...
** package 'praise' successfully unpacked and MD5 sums checked
** using staged installation
** R
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** DONE (praise)
```

###

`install.packages()` is base R's package manager: it downloads one package from CRAN and compiles or unpacks it into your library, one call at a time. Contrast this with **pak**: pak resolves and installs entire dependency trees in parallel, which is why the image build (and the dormant refresh recipe you read in Exercise 16) uses `pak::pkg_install()` for the four course packages.

### Exercise 18

In the bash Terminal, run `/opt/venv/bin/pip install emoji`. CP/CR.

```{r}
#| label: codespace-starter-18
#| echo: false
learnr2::question(
  "In the bash Terminal, run /opt/venv/bin/pip install emoji. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
codespace-starter $ /opt/venv/bin/pip install emoji
Collecting emoji
  Downloading emoji-2.14.0-py3-none-any.whl (590 kB)
Installing collected packages: emoji
Successfully installed emoji-2.14.0
```

###

`/opt/venv` is a **virtual environment**: a self-contained folder of Python packages --- numpy, pandas, matplotlib, and more --- baked into the image and kept separate from the system Python. It is also first on your `PATH`, so plain `pip` already resolves to `/opt/venv/bin/pip`; naming the full path just makes the target explicit, a habit that pays off on machines with more than one Python. Either way the package lands in the environment `quarto render` uses for Python chunks, so `import emoji` works immediately.

### Exercise 19

Ask AI to install the R package `cowsay` for you, and to tell you exactly what command it ran. CP/CR.

```{r}
#| label: codespace-starter-19
#| echo: false
learnr2::question(
  "Ask AI to install the R package cowsay for you, and to tell you exactly what command it ran. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
> install.packages("cowsay")
Installing package into '/home/ppbds-student/R/library'
(as 'lib' is unspecified)
trying URL 'https://cloud.r-project.org/src/contrib/cowsay_0.9.0.tar.gz'
downloaded 22 KB

* installing *source* package 'cowsay' ...
** package 'cowsay' successfully unpacked and MD5 sums checked
** using staged installation
** R
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** DONE (cowsay)
```

###

Whether you type `install.packages()` yourself or ask AI to install cowsay, the same function ultimately runs underneath --- an AI agent calls the same R functions and Terminal commands you would type by hand. That's why it's worth reading the exact command AI reports back, the same way you read your own in Exercise 17.

### Exercise 20

Ask AI to install the Python library `pyfiglet` into this Codespace's Python environment, and to tell you exactly what command it ran. CP/CR.

```{r}
#| label: codespace-starter-20
#| echo: false
learnr2::question(
  "Ask AI to install the Python library pyfiglet into this Codespace's Python environment, and to tell you exactly what command it ran. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
codespace-starter $ /opt/venv/bin/pip install pyfiglet
Collecting pyfiglet
  Downloading pyfiglet-1.0.2-py3-none-any.whl (1.1 MB)
Installing collected packages: pyfiglet
Successfully installed pyfiglet-1.0.2
```

###

A good AI agent in this Codespace targets `/opt/venv/bin/pip` --- the environment `quarto render` uses for Python chunks (Exercise 18) --- rather than guessing. An assistant without that context might install into some other Python and leave `import pyfiglet` failing silently later. Checking which `pip` actually ran is worth doing before trusting the result, not after.

## Summary
###

This tutorial toured the `devcontainer.json` behind `codespace-starter` itself: its pinned custom PPBDS image, its VS Code extensions array, the settings that control the R Terminal, plot rendering, Git commits, autosave, word wrap, and tab behavior, the `onCreateCommand` behind the ready banner, and the retired `postCreateCommand` --- course packages now ship baked into the image, with the refresh recipe kept dormant in a comment. The next tutorial, "The Devcontainer Dockerfile," goes one level deeper and reads through the Dockerfile that actually builds the `ghcr.io/ppbds/devcontainer` image this `devcontainer.json` points at.

### Exercise 1

In the bash Terminal, in your `our-codespace-starter` repo, run `git status`. CP/CR.

```{r}
#| label: summary-1
#| echo: false
learnr2::question(
  "In the bash Terminal, in your our-codespace-starter repo, run git status. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
our-codespace-starter $ git status
On branch main
nothing to commit, working tree clean
our-codespace-starter $
```

###

A clean `git status` here just confirms Git isn't tracking anything you haven't dealt with --- this tutorial never asked you to change files in your own repo, only to read `codespace-starter`'s.

### Exercise 2

Paste the URL of your `our-codespace-starter` repository.

```{r}
#| label: summary-2
#| echo: false
learnr2::question(
  "Paste the URL of your our-codespace-starter repository.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
https://github.com/ppbds-student/our-codespace-starter
```

###

This repo now exists on GitHub independent of any one Codespace --- you or a grader can reconnect to it and pick your work back up even after the Codespace itself is gone.

### Exercise 3

Stop this Codespace: open the Command Palette with `Ctrl+Shift+P` (or `Cmd+Shift+P` on Mac), type `>stop`, and select **"Codespaces: Stop Current Codespace"**. Then go to https://github.com/codespaces and paste the row for it, showing its "Stopped" status.

```{r}
#| label: summary-3
#| echo: false
learnr2::question(
  "Stop this Codespace: open the Command Palette with Ctrl+Shift+P (or Cmd+Shift+P on Mac), type >stop, and select \"Codespaces: Stop Current Codespace\". Then go to https://github.com/codespaces and paste the row for it, showing its \"Stopped\" status.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
our-codespace-starter    ppbds-student/our-codespace-starter    main    Stopped
```

###

Stopped Codespaces don't use compute hours but still occupy storage, so they can be restarted any time --- deleting one instead frees the storage but can't be undone.

## Download answers

```{r}
#| label: download-answers-1
#| echo: false
learnr2::question(
  "How many minutes, approximately, did it take you to complete this
  tutorial? For example, an hour and a half would be 90 minutes.",
  type = "reflection_editable",
  validate = "integer"
)
```

###

```{r}
#| label: download-answers-2
#| echo: false
learnr2::download_answers_button(filename_prefix = "our-codespace-starter")
```
