---
title: "The Devcontainer Dockerfile"
author:
  - David Kane
  - Krrish Jayakumar
tutorial:
  id: "12-docker"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "A reading tour of the real Dockerfile that builds ghcr.io/ppbds/devcontainer --- the image codespace-starter's devcontainer.json points at. Covers digest pinning, layer ordering, system libraries, permissions, R and Python package installs, and the smoke tests that guard the build."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In `11-codebase-start` you found the line `"image": "ghcr.io/ppbds/devcontainer:1.0.1"` inside `codespace-starter`'s `devcontainer.json`. That image does not appear out of nowhere --- it is built from a Dockerfile maintained by PPBDS. This tutorial reads through that real, current Dockerfile, section by section, to understand what it does and why it is written the way it is.

###

This is a **reading tour, not a building exercise**. You will not build this image yourself --- it is not buildable inside the course Codespace, and doing so is not the point. Instead, each exercise shows you a piece of the Dockerfile and asks you to find and quote a specific line or block from it, then explains what that piece does and why it exists.

## The Devcontainer Image
###

Here is the complete Dockerfile that builds `ghcr.io/ppbds/devcontainer`. Skim it once, then work through the exercises below, which walk through it piece by piece, in the order the layers actually appear.

````
# syntax=docker/dockerfile:1.7
#
# THE PPBDS devcontainer image — ghcr.io/ppbds/devcontainer.
#
# One image for everyone: students taking the course (launched via
# PPBDS/codespace-starter) AND developers working on PPBDS packages
# (PPBDS/primer etc.). It replaced the old base/dev/student trio in v1.0.0 —
# the dev image's delta (qpdf + five R packages) was a rounding error against
# the full image, and the split cost more in CI plumbing and cross-image
# permission bugs than it saved. The old images remain on GHCR at their final
# tags but receive no new versions.
#
# Layer order matters and mirrors the old base → dev → student layering:
# system libs and CLIs first (root), R stacks next (root, then hardened),
# then course packages and per-user installs (rstudio), then Python/JS
# (root), then the end-to-end smoke tests.

# Pinned by digest. The :4.5 tag floats as Rocker patches the image; the
# digest freezes us at a known build. To bump: pull :4.5 fresh, copy the
# new digest in, commit deliberately. Do not switch to :4.5 unpinned.
# R version note: this base is R 4.5 (rocker's newest devcontainer image).
# R 4.6.0 shipped 2026-04-24 but rocker has not yet published a :4.6
# devcontainer image; revisit once it exists and its P3M binaries settle.
FROM ghcr.io/rocker-org/devcontainer/tidyverse:4.5@sha256:289c5d02d8115aa209f4a8a49ee9378dccbf623897eed9cc46c87dfbbca9015b

ARG QUARTO_VERSION=1.9.38
ARG ARF_VERSION=0.3.4
ARG NODE_MAJOR=22

# AI CLI versions. Pinned so builds are reproducible and the baked version
# doesn't silently drift behind the registry. Bump deliberately, like
# Quarto/arf. (Antigravity CLI is the exception — its installer offers no
# version pin, so `agy` floats; see the agy install block below.)
ARG CLAUDE_CODE_VERSION=2.1.181
ARG CODEX_VERSION=0.141.0

# System libraries needed by the R stacks below.
#  - Geospatial: sf, terra, etc.
#  - text shaping libs: ragg / textshaping (modern ggplot2 graphics)
#  - cmake: build tool for source packages like fs (pulled in by primer.tutorials)
#  - libuv1-dev: fs >= 2.1.0 declares libuv as a sysreq. The P3M binary works
#    without it, but pak prints a phantom "✖ Missing 1 system package:
#    libuv1-dev" on every install that touches fs (i.e. nearly all of them) —
#    same annoyance class as the libnode-dev story below, solved here by just
#    baking the real (small) lib.
#  - qpdf: required by `R CMD check --as-cran` for PDF manual checks (package
#    development; rocker/tidyverse already provides the rest of the
#    build/check toolchain).
# NOTE: libv8/libnode-dev is deliberately NOT installed here. The V8 R package
# binary from Posit Package Manager is statically linked (bundled libv8) and
# needs no system library, AND NodeSource's nodejs (installed below) ships
# overlapping files that would remove a distro libnode-dev anyway. pak is told
# the requirement is satisfied via the metadata-only libnode-dev-virtual package
# further down — see that block for the full rationale.
RUN apt-get update && apt-get install -y --no-install-recommends \
        libgdal-dev \
        libproj-dev \
        libgeos-dev \
        libudunits2-dev \
        libfontconfig1-dev \
        libharfbuzz-dev \
        libfribidi-dev \
        cmake \
        libuv1-dev \
        qpdf \
    && rm -rf /var/lib/apt/lists/*

# GitHub CLI. Official install method per the gh docs.
RUN curl -fsSL https://cli.github.com/packages/githubcli-archive-keyring.gpg \
        -o /usr/share/keyrings/githubcli-archive-keyring.gpg \
    && chmod go+r /usr/share/keyrings/githubcli-archive-keyring.gpg \
    && echo "deb [arch=$(dpkg --print-architecture) signed-by=/usr/share/keyrings/githubcli-archive-keyring.gpg] https://cli.github.com/packages stable main" \
        > /etc/apt/sources.list.d/github-cli.list \
    && apt-get update && apt-get install -y --no-install-recommends gh \
    && rm -rf /var/lib/apt/lists/*

# AI coding-assistant CLIs. Four tools spanning several providers and billing
# models (aider is itself multi-provider). Students choose which to use. The
# recommended way to authenticate is to SIGN IN on first run with the matching
# account (a flat-rate plan, not metered API calls); an API key supplied via
# Codespaces user secrets (https://github.com/settings/codespaces) is the
# fallback:
#   - claude  (Claude Code, Anthropic)  — sign in with a Claude plan, or ANTHROPIC_API_KEY
#   - codex   (Codex CLI, OpenAI)        — sign in with a ChatGPT plan, or `codex login`
#   - agy     (Antigravity CLI, Google)  — sign in with a Google account, or ANTIGRAVITY_API_KEY
#   - aider   (multi-provider)           — key-only: DeepSeek/OpenRouter/OpenAI/Anthropic
#
# NOTE: Google's Gemini CLI was REMOVED. Google EOL'd the consumer Gemini CLI
# on 2026-06-18 — the free/sign-in path stopped serving requests — and pushed
# everyone to Antigravity. The Google slot is now the Antigravity CLI (`agy`,
# installed further below via Google's curl installer, alongside arf).
#
# Node hosts the npm-distributed CLIs (claude, codex). We install from
# NodeSource rather than apt: Ubuntu Noble ships Node 18.x, which is past end of
# life — we want a current Node LTS for the npm CLIs. NodeSource gives us one
# (pinned via NODE_MAJOR above). pipx installs aider into an isolated venv with
# shims in /usr/local/bin (on every user's PATH), avoiding the PEP 668 lockout
# on Noble's system Python.
RUN curl -fsSL "https://deb.nodesource.com/setup_${NODE_MAJOR}.x" | bash - \
    && apt-get install -y --no-install-recommends \
        nodejs \
        pipx \
    && rm -rf /var/lib/apt/lists/*

# Tell pak that the V8 R package's system requirement is satisfied. pak's
# sysreqs database maps V8 to the Debian package "libnode-dev", but:
#   (1) the V8 binary we get from Posit Package Manager is statically linked
#       (bundled libv8) and needs no system library at all, and
#   (2) NodeSource's nodejs (installed just above) ships overlapping files and
#       removes any distro libnode-dev, so it can't be present anyway.
# Without this, pak prints "✖ Missing 1 system package: libnode-dev" on every
# student install that pulls in V8 (gt, primer.tutorials, …) — alarming noise
# for beginners, even though nothing is actually missing. A metadata-only
# package that Provides libnode-dev/libv8-dev satisfies pak's check the
# canonical Debian way; it ships no files, so it coexists with NodeSource
# nodejs. (Validated against student:0.6.1 to silence the message without
# changing what is actually linked.) The trailing dpkg-query is a build-time
# smoke test: fail the build if the Provides did not register.
RUN mkdir -p /tmp/libnode-dev-virtual/DEBIAN \
    && printf '%s\n' \
        'Package: libnode-dev-virtual' \
        'Version: 1.0' \
        'Architecture: all' \
        'Maintainer: PPBDS <noreply@ppbds.invalid>' \
        'Provides: libnode-dev, libv8-dev' \
        'Description: Virtual provide for libnode-dev / libv8-dev' \
        ' The V8 R package ships a bundled static libv8 and needs no system' \
        ' library; this satisfies pak sysreqs without NodeSource-conflicting pkgs.' \
        > /tmp/libnode-dev-virtual/DEBIAN/control \
    && dpkg-deb --build /tmp/libnode-dev-virtual /tmp/libnode-dev-virtual.deb \
    && dpkg -i /tmp/libnode-dev-virtual.deb \
    && rm -rf /tmp/libnode-dev-virtual /tmp/libnode-dev-virtual.deb \
    && dpkg-query -W -f='${Provides}\n' libnode-dev-virtual | grep -q libnode-dev

ENV PIPX_HOME=/opt/pipx \
    PIPX_BIN_DIR=/usr/local/bin

RUN npm install -g \
        "@anthropic-ai/claude-code@${CLAUDE_CODE_VERSION}" \
        "@openai/codex@${CODEX_VERSION}" \
    && pipx install aider-chat

# Pre-seed Codex CLI config for the runtime user (rstudio). The one setting
# that matters for an immutable image:
#   - check_for_update_on_startup=false: the CLI is a global npm install the
#     runtime user can't overwrite, so a startup update check is pointless
#     noise. The Codex docs say to disable it "when updates are centrally
#     managed" — that's us; CODEX_VERSION above is the deliberate version knob.
# Onboarding tooltips and analytics are left at their defaults. Codex rewrites
# this file as students use it, so nothing here is locked.
RUN install -d -o rstudio -g rstudio /home/rstudio/.codex \
    && printf '%s\n' 'check_for_update_on_startup = false' \
        > /home/rstudio/.codex/config.toml \
    && chown rstudio:rstudio /home/rstudio/.codex/config.toml

# Quarto. Use dpkg --print-architecture so the same Dockerfile works on
# amd64 and arm64 builders.
RUN set -eux; \
    arch="$(dpkg --print-architecture)"; \
    curl -fsSL -o /tmp/quarto.deb \
        "https://github.com/quarto-dev/quarto-cli/releases/download/v${QUARTO_VERSION}/quarto-${QUARTO_VERSION}-linux-${arch}.deb"; \
    dpkg -i /tmp/quarto.deb; \
    rm /tmp/quarto.deb; \
    quarto --version

# Per-user installers, run as rstudio so their binaries land under
# /home/rstudio (arf in ~/.cargo/bin, agy in ~/.local/bin):
#  - arf: Rust-based R console. Path matches what consumer devcontainer.json
#    files reference as r.rterm.linux.
#  - agy: Antigravity CLI, Google's terminal coding agent and the successor to
#    the now-EOL'd Gemini CLI. NOTE: its installer offers NO version pin (always
#    latest), so unlike every other tool here agy floats with each image
#    rebuild. Accepted because there is no pin mechanism; the CLI smoke test
#    below still fails the build if a future release breaks. Auth at runtime is
#    Google sign-in (prints a URL + one-time code, works headless in Codespaces)
#    or an ANTIGRAVITY_API_KEY from Google AI Studio.
USER rstudio
RUN curl --proto '=https' --tlsv1.2 -fsSL \
        "https://github.com/eitsupi/arf/releases/download/v${ARF_VERSION}/arf-console-installer.sh" | sh
RUN curl -fsSL https://antigravity.google/cli/install.sh | bash
USER root

# Symlink both per-user binaries to a stable system path so consumer config
# doesn't hard-code "/home/rstudio/...". If the image's default user ever
# changes, only this Dockerfile needs updating, not every consumer.
RUN ln -s /home/rstudio/.cargo/bin/arf /usr/local/bin/arf \
    && ln -s /home/rstudio/.local/bin/agy /usr/local/bin/agy

# pak: fast parallel R package installer, used for every R install below.
RUN R -q -e 'install.packages("pak", repos = sprintf("https://r-lib.github.io/p/pak/stable/%s/%s/%s", .Platform$pkgType, R.Version()$os, R.Version()$arch))'

# tidymodels + a set of modeling engines. All of these are pulled from the
# CURRENT P3M snapshot (overriding rocker's frozen one) so we get versions
# recent enough for the CatBoost engine (parsnip >= 1.4.0, bonsai >= 0.4.1)
# and current everything else. None of the engine packages ship inside the
# tidymodels meta-package — parsnip only provides the interface, so each
# backend is installed explicitly here:
#   - tidymodels  : the modeling framework (parsnip/recipes/rsample/tune/yardstick/…)
#   - bonsai      : parsnip bridge for the lightgbm AND catboost engines
#   - xgboost, lightgbm, randomForest, ranger, glmnet : tree/regularized engines
#   - brms        : Bayesian regression via Stan — heavy (pulls rstan/StanHeaders);
#                   compiles models at runtime using the rocker C++ toolchain
#   - BH, RcppEigen, RcppParallel : the Boost / Eigen / TBB C++ headers rstan
#   - remotes     : needed for the CatBoost URL install below
# CatBoost is not on CRAN — install its pinned linux-x86_64 release binary via
# remotes::install_url; the INSTALL_opts are what a clean binary install needs
# (per the catboost docs / the 2026-06 Posit tidymodels+catboost blog). The image
# is amd64-only (build.yml sets no platforms), so the x86_64 binary suffices. The
# end-to-end fit smoke test near the bottom proves the engine actually works —
# important, since we install it with --no-test-load.
RUN R -q -e 'options(repos = c(P3M = "https://packagemanager.posit.co/cran/__linux__/noble/latest")); pak::pkg_install(c("tidymodels", "bonsai", "remotes", "xgboost", "lightgbm", "randomForest", "ranger", "glmnet", "brms", "BH", "RcppEigen", "RcppParallel"))' \
 && R -q -e 'remotes::install_url("https://github.com/catboost/catboost/releases/download/v1.2.10/catboost-R-linux-x86_64-1.2.10.tgz", INSTALL_opts = c("--no-multiarch", "--no-test-load", "--no-staged-install"))'

# Inference-reporting + presentation packages. These are used pervasively in
# BOTH PPBDS/primer's book/ and its tutorials (audited 2026-07), but pak never
# installed them because primer.tutorials only lists them as Suggests —
# students hit an install wall at the first library() call:
#   - gt              : presentation-quality tables (1,100+ gt:: calls in the
#                       book; was previously present only transitively via
#                       primer.tutorials' Imports — now a deliberate choice)
#   - marginaleffects : predictions/comparisons/slopes from fitted models —
#                       the book's standard post-estimation workflow
#   - patchwork       : ggplot composition ("p1 + p2")
#   - easystats       : meta-package (parameters/performance/effectsize/see/…)
#                       used across the tutorials
RUN R -q -e 'options(repos = c(P3M = "https://packagemanager.posit.co/cran/__linux__/noble/latest")); pak::pkg_install(c("gt", "marginaleffects", "patchwork", "easystats"))'

# httpgd (graphics device for the VS Code R extension) is intentionally
# NOT installed here: the rocker devcontainer base already bakes it in via
# its r-packages feature ("packages": "httpgd"), so a second install was
# pure redundancy. Do not re-add it. The smoke test at the bottom of this
# Dockerfile asserts httpgd loads, so if a future base image ever drops
# it, the build fails loudly here instead of students hitting a missing
# graphics device at runtime.

# Headless-container workarounds: any consumer (Codespaces, local Docker,
# JetBrains Gateway) lacks a desktop, so anything that tries to open a
# browser fails. Stub xdg-open and set BROWSER to a no-op.
RUN printf '#!/bin/sh\nexit 0\n' > /usr/local/bin/xdg-open \
    && chmod +x /usr/local/bin/xdg-open
ENV BROWSER=/usr/bin/true

# All R installs above ran as root, so the resulting package dirs are
# root-owned with group-read-only permissions. Restore rocker's convention
# of group-writable site-library so the runtime user (rstudio, in staff
# group) can install/update packages — and so package-load paths that
# touch lock files or caches don't trip permission errors.
#
# Two hardenings, both for the same failure class (a root pak run poisons
# the library for later non-root installs; see the rstudio-user rationale
# below):
#  - rm -rf _cache: pkgdepends locks site-library/_cache/<pkg>.lock during
#    installs. Locks created by ROOT here are group-unreadable (root's
#    umask), and a later rstudio-user pak run then dies with
#    "filelock::lock(): Cannot open lock file: Permission denied" — this
#    broke the student build (2026-07). The dir is pure scratch; delete it.
#  - g+rwX (not just g+w): write alone leaves group-READ missing on any
#    file root created without it, and no traverse on such dirs.
RUN rm -rf /usr/local/lib/R/site-library/_cache \
    && chmod -R g+rwX /usr/local/lib/R/site-library

# Never let pak try to `sudo apt-get install` system requirements. The system
# libraries our packages need are baked above; pak running as the non-root
# rstudio user can't sudo, and (e.g. for V8's libnode-dev) it even mis-detects
# baked libs as missing and tries anyway, which fails the build / postCreate /
# a student's runtime install. With this off pak still PRINTS any sysreqs but
# builds against the baked libs. (Add new system deps to the apt block at the
# top, not via pak.) Applies to the rstudio-user installs below, to
# codespace-starter's postCreateCommand, and to runtime installs.
ENV PKG_SYSREQS=false

# ── Everything below installs as rstudio, not root ───────────────────────────
# This is the design fix for a class of permissions bugs: when pak runs as
# root, it drops lock files in site-library as -rw--w---- (group write but no
# read, due to root's restrictive umask), which then blocks every subsequent
# install by the non-root rstudio user with "filelock::lock(): Permission
# denied". Running as rstudio in the first place means the lock files get a
# normal umask and never poison the library for later installs.
#
# This relies on rocker's site-library being drwxrwsr-x with group=staff and
# rstudio in the staff group (plus the g+rwX hardening above).
USER rstudio

# Package-development R packages (the old dev image's payload). No
# `upgrade = TRUE` here — these have stable, well-behaved binary builds in
# P3M and haven't shown the ABI-mismatch failure mode in practice. The smoke
# test below fails the build if that ever changes.
RUN R -q -e 'pak::pkg_install(c("devtools", "pkgdown", "roxygen2", "testthat", "usethis"))' \
    && R --vanilla -e 'for (p in c("devtools", "pkgdown", "roxygen2", "testthat", "usethis")) if (!requireNamespace(p, quietly = TRUE)) stop("smoke test failed to load: ", p)'

# Course R packages, all installed from GitHub source via pak so a fresh
# image always tracks the latest commit on each package's default branch.
# This matches PPBDS practice — the rest of the PPBDS ecosystem expects
# the development version, not whatever an r-universe build cycle (or
# CRAN) most recently blessed.
#
# These four are also refreshed to the latest commit on every Codespace
# create via codespace-starter's postCreateCommand; baking them here pre-
# installs the whole dependency tree (so that refresh is a quick update, not
# a cold build) and leaves a working fallback if GitHub is unreachable. Note
# primer.tutorials was SPLIT OUT of the primer repo (2026-07): as a subdir
# install ("PPBDS/primer/primer.tutorials") every install downloaded the
# full ~183 MB primer tarball — book, class exercises — to deliver a ~4 MB
# package, at every image build AND every Codespace create. Do not point
# this back at the subdir path. Add further "PPBDS/<name>" entries
# (in BOTH places) when the syllabus needs them.
#
# upgrade = TRUE forces pak to pull the latest version of every
# transitive dep (learnr, knitr, rmarkdown, xfun, ...). Without this, a
# binary dep built against a newer version of xfun than what rocker
# shipped fails to load with "object 'attr' is not exported by
# 'namespace:xfun'".
RUN R -q -e 'pak::pkg_install(c( \
        "PPBDS/tutorial.helpers", \
        "PPBDS/vscode.tutorials", \
        "PPBDS/misc.tutorials", \
        "PPBDS/primer.tutorials" \
    ), upgrade = TRUE)'

# Smoke test 1: every baked-in package and the learnr/knitr/rmarkdown
# chain must all load. The original learnr/xfun ABI mismatch failure
# would have been caught here at build time.
RUN R --vanilla -e 'for (p in c("tutorial.helpers", "vscode.tutorials", "misc.tutorials", "primer.tutorials", "learnr", "knitr", "rmarkdown")) if (!requireNamespace(p, quietly = TRUE)) stop("smoke test failed to load: ", p)'

# Smoke test 2: rstudio can install a fresh package into site-library
# without permission errors. Catches the "root-built image leaves
# unwritable lock files" regression at build time instead of at
# first-student-pak-call time. praise is tiny (~5KB) and dependency-free.
RUN R --vanilla -e 'install.packages("praise", repos = "https://cloud.r-project.org"); if (!requireNamespace("praise", quietly = TRUE)) stop("rstudio cannot install into site-library — check site-library permissions above")' \
 && R --vanilla -e 'remove.packages("praise")'

# R interactive-visualisation + Shinylive packages — htmlwidgets for "fancy"
# interactive output that still publishes to STATIC hosting (GitHub Pages):
# plotly + leaflet (charts/maps), DT (interactive tables), crosstalk (client-
# side linking/filtering, no server); plus shiny + shinylive, which compile a
# Shiny app to WebAssembly (webR) so it too runs on a static host. pak via P3M
# Linux binaries — fast (~16 s, ~70 MB). (Python equivalents — plotly/altair/
# folium/itables/shiny/shinylive — are in requirements.lock.)
RUN R -q -e 'pak::pkg_install(c("plotly", "leaflet", "DT", "crosstalk", "shiny", "shinylive"))' \
    && R --vanilla -e 'for (p in c("plotly","leaflet","DT","crosstalk","shiny","shinylive")) if (!requireNamespace(p, quietly = TRUE)) stop("smoke test failed to load: ", p)'

# Mapping / census stack — NYT-style tract maps (dot-density, choropleths):
# sf (vector geodata; links against the GDAL/PROJ/GEOS/udunits system libs
# baked above) and tidycensus (ACS/decennial data + geometry in one call;
# pulls in tigris for TIGER shapefiles). Students need a free Census API key
# for live tidycensus queries. P3M Linux binaries, so this is fast. The
# interactive path needs nothing more: leaflet (above) draws sf objects, and
# free CARTO basemap tiles substitute for Mapbox.
#
# Plus the rest of the Kyle Walker toolkit:
#  - mapgl:     MapLibre/Mapbox GL htmlwidgets — WebGL vector maps (the NYT
#               look) with NO token via maplibre() + CARTO styles; publishes
#               static to GitHub Pages like every other htmlwidget here.
#  - crsuggest: suggests the right projected CRS for a dataset — the one-
#               function answer to the #1 beginner mapping confusion.
#  - idbr:      Census International Data Base (population pyramids, country
#               time series); uses the same Census API key as tidycensus.
# Deliberately NOT mapboxapi: it requires a per-student Mapbox account/token.
RUN R -q -e 'pak::pkg_install(c("sf", "tidycensus", "mapgl", "crsuggest", "idbr"))' \
    && R --vanilla -e 'for (p in c("sf","tidycensus","tigris","mapgl","crsuggest","idbr")) if (!requireNamespace(p, quietly = TRUE)) stop("smoke test failed to load: ", p)' \
    && R --vanilla -e 'library(sf); p <- st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,1), c(0,0)))), crs = 4326); pts <- st_sample(p, 10); stopifnot(length(pts) == 10); cat("sf geometry ops OK\n")'

USER root

# ── Python data-science stack ────────────────────────────────────────────────
# Lets students do data science in Python too, not just R: numpy/pandas
# (wrangling), matplotlib/seaborn (plots), scikit-learn (ML), statsmodels
# (regression/inference — the lm() analog), and jupyter/ipykernel (notebooks;
# also what Quarto's Python engine and the VS Code Jupyter extension drive).
# We pin the full dependency tree in requirements.lock (generated from
# requirements.txt via `uv pip compile`); edit the .txt and recompile to
# bump versions.
#
# Installed with uv (Astral's fast resolver/installer) into a venv at /opt/venv:
#   --seed        → put pip/setuptools/wheel IN the venv, so students can plain
#                   `pip install` more packages (uv venvs omit pip by default).
#   --python /usr/bin/python3 → use the system CPython, NOT a uv-managed
#                   download, so it's a normal, predictable interpreter.
#   chgrp staff + g+w → make the venv group-writable (rstudio is in staff), the
#                   same treatment R's site-library gets, so a student's
#                   `pip install` actually has somewhere to write.
# /opt/venv goes first on PATH, so `python`/`pip`/`jupyter` are the venv's.
# Quarto runs Python via the jupyter engine (ipykernel) — no reticulate needed.
# We still set RETICULATE_PYTHON so that IF someone installs R's reticulate
# (not baked in) it bridges to this same venv rather than a separate Python.
# uv's own binary is copied from its official pinned image (pinned, unlike agy).
COPY --from=ghcr.io/astral-sh/uv:0.11.23 /uv /usr/local/bin/uv
COPY requirements.lock /tmp/requirements.lock
COPY requirements.txt /tmp/requirements.txt

# Lock-staleness guard: every top-level package in requirements.txt must appear
# in the lock, so editing the .txt and forgetting to re-run `uv pip compile`
# fails the build instead of silently shipping a stale lock. Presence-only (no
# version comparison), so it never false-fails when PyPI publishes new
# transitive versions.
RUN set -eu; \
    while IFS= read -r line; do \
        case "$line" in ''|\#*) continue ;; esac; \
        pkg="${line%%[=<>!~ ]*}"; \
        grep -qiE "^${pkg}==" /tmp/requirements.lock \
          || { echo "requirements.lock is STALE: '${pkg}' is in requirements.txt but not the lock — run: uv pip compile requirements.txt -o requirements.lock" >&2; exit 1; }; \
    done < /tmp/requirements.txt

RUN uv venv --seed --python /usr/bin/python3 /opt/venv \
    && uv pip install --python /opt/venv/bin/python --no-cache -r /tmp/requirements.lock \
    && /opt/venv/bin/python -m ipykernel install --prefix /usr/local \
         --name python3 --display-name "Python (data science)" \
    && chgrp -R staff /opt/venv \
    && chmod -R g+w /opt/venv \
    && rm /tmp/requirements.lock /tmp/requirements.txt
ENV PATH=/opt/venv/bin:$PATH \
    RETICULATE_PYTHON=/opt/venv/bin/python

# Smoke test 1: the whole stack must import (catches a bad lock / ABI break at
# build time instead of at a student's first import).
RUN python -c "import numpy, pandas, matplotlib, seaborn, sklearn, statsmodels, ipykernel, plotly, altair, folium, itables, shiny, shinylive; print('py-ds stack OK')"

# Smoke test 2: a non-root student (rstudio, in staff) can pip-install into the
# venv — the Python analog of the R 'praise' test above. --seed gave the venv a
# pip; the group-writable perms let rstudio use it.
RUN su rstudio -c "/opt/venv/bin/pip install --no-cache-dir --quiet cowsay" \
    && /opt/venv/bin/python -c "import cowsay; print('rstudio-can-install OK')" \
    && /opt/venv/bin/pip uninstall -y --quiet cowsay
# ─────────────────────────────────────────────────────────────────────────────

# ── Observable Framework ─────────────────────────────────────────────────────
# Observable's open-source static-site generator for interactive data apps and
# dashboards (JavaScript front-end + any-language back-end). Adds the
# `observable` CLI (`observable create`, `observable preview`, `observable
# build`). Node is already present (installed above for the AI CLIs); ~84 MB.
#
# NOTE: interactive Observable charts INSIDE a Quarto document need NOTHING
# extra — Quarto's bundled `{ojs}` cells already render Observable Plot / OJS
# (build-test verified). Framework is only for standalone data-app projects.
# The `observable --version` call is the smoke test (fails the build if broken).
RUN npm install -g @observablehq/framework@1.13.4 && observable --version
# ─────────────────────────────────────────────────────────────────────────────

# ── Image-wide smoke tests ───────────────────────────────────────────────────
# R smoke test: foundational packages must load. Catches binary-ABI
# mismatches at build time instead of at Codespace-launch time.
# --vanilla skips Rprofile so this isolates the package-load path itself.
RUN R --vanilla -e 'for (p in c("pak", "httpgd")) if (!requireNamespace(p, quietly = TRUE)) stop("smoke test failed to load: ", p)'

# Modeling smoke test 1: tidymodels, bonsai, every engine, and brms must load.
RUN R --vanilla -e 'for (p in c("tidymodels", "bonsai", "xgboost", "lightgbm", "randomForest", "ranger", "glmnet", "brms", "catboost")) if (!requireNamespace(p, quietly = TRUE)) stop("modeling smoke test failed to load: ", p)'

# Inference/presentation smoke test: the primer book+tutorial packages must load.
RUN R --vanilla -e 'for (p in c("gt", "marginaleffects", "patchwork", "easystats")) if (!requireNamespace(p, quietly = TRUE)) stop("inference/presentation smoke test failed to load: ", p)'

# Modeling smoke test 2: CatBoost end-to-end THROUGH tidymodels/bonsai — fit a
# tiny model and predict, not just load the package. Since catboost is installed
# with --no-test-load, this is the real proof the binary + engine binding work.
RUN R --vanilla -e 'suppressPackageStartupMessages({library(tidymodels); library(bonsai)}); m <- boost_tree(trees = 5, mode = "regression") |> set_engine("catboost") |> fit(mpg ~ ., data = mtcars); stopifnot(nrow(predict(m, mtcars)) == nrow(mtcars)); cat("catboost + bonsai end-to-end OK\n")'

# Modeling smoke test 3: brms end-to-end — fit a trivial model so the build
# actually COMPILES a Stan program with the image's toolchain. A load check
# wouldn't catch a broken Stan/compiler config; this does. chains/iter are tiny
# since we only care that compilation + sampling run, not about the estimates.
RUN R --vanilla -e 'fit <- brms::brm(mpg ~ wt, data = mtcars, chains = 1, iter = 100, refresh = 0, silent = 2); stopifnot(inherits(fit, "brmsfit")); cat("brms + Stan end-to-end OK\n")'

# CLI smoke test: every shell tool must be on PATH and runnable.
RUN claude --version && codex --version && agy --version && aider --version

# ── End-to-end authoring smoke tests ─────────────────────────────────────────
# These run last, once R + Python + Quarto are all in place, so they exercise
# the actual student authoring path (not just that packages load). We render an
# R doc and a Python doc SEPARATELY — that's how students work (one language per
# doc), and it covers BOTH Quarto engines without needing reticulate: an R doc
# uses the knitr engine, a Python-only doc uses the jupyter engine (the venv's
# ipykernel). NOTE: these are build-time and headless, so they cannot catch VS
# Code / extension runtime issues (e.g. the Python extension activating the venv
# in the R terminal) — that class needs a manual "launch a Codespace and run a
# tutorial" check.
#
# (a) An R-only Quarto doc must render to HTML (knitr engine).
RUN set -eux; \
    d="$(mktemp -d)"; cd "$d"; \
    printf '%s\n' '---' 'title: r' 'format: html' '---' '' \
      '```{r}' 'summary(cars$speed)' '```' > r.qmd; \
    quarto render r.qmd --to html; test -s r.html; echo "QUARTO-R OK"; \
    cd /; rm -rf "$d"

# (b) A Python-only Quarto doc must render to HTML (jupyter engine → venv kernel).
RUN set -eux; \
    d="$(mktemp -d)"; cd "$d"; \
    printf '%s\n' '---' 'title: py' 'format: html' 'jupyter: python3' '---' '' \
      '```{python}' 'import pandas as pd' 'pd.DataFrame({"a": [1, 2]}).describe()' '```' > p.qmd; \
    quarto render p.qmd --to html; test -s p.html; echo "QUARTO-PY OK"; \
    cd /; rm -rf "$d"

# (c) The course tutorials must be discoverable (catches a broken tutorial pkg).
RUN R --vanilla -e 'ts <- learnr::available_tutorials("tutorial.helpers"); if (!("getting-started" %in% ts$name)) stop("getting-started tutorial not found"); cat("TUTORIALS OK:", paste(ts$name, collapse = ", "), "\n")'
# ─────────────────────────────────────────────────────────────────────────────

# Locale (en_US.UTF-8) and timezone (Etc/UTC) are inherited from the rocker base.
````

### Exercise 1

Find the `FROM` line near the top of the Dockerfile. Copy and paste it.

```{r docker-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
FROM ghcr.io/rocker-org/devcontainer/tidyverse:4.5@sha256:289c5d02d8115aa209f4a8a49ee9378dccbf623897eed9cc46c87dfbbca9015b
```

###

The image is pinned by **digest** (the long `sha256:...` hash), not just by the `:4.5` tag. A tag like `:4.5` floats --- Rocker can push a new build under that same tag at any time, silently changing what every future build of this Dockerfile pulls in. The digest freezes the base at one exact, immutable build. Bumping it is a deliberate act: pull `:4.5` fresh, copy in the new digest, and commit the change on purpose, rather than being surprised by an upstream update.

### Exercise 2

Find the sentence in the header comment that explains why the old base/dev/student image trio was removed. Copy and paste it.

```{r docker-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look like:

```
It replaced the old base/dev/student trio in v1.0.0 —
the dev image's delta (qpdf + five R packages) was a rounding error against
the full image, and the split cost more in CI plumbing and cross-image
permission bugs than it saved.
```

###

There is now **one image for everyone**: students launched via `codespace-starter` and PPBDS package developers alike. The old three-image split existed to keep the student image lean, but the actual difference between the "dev" and "student" images was tiny (qpdf plus five R packages) --- not worth the extra CI configuration and the permission bugs that came from images built slightly differently from each other.

### Exercise 3

Find the comment describing the Dockerfile's overall layer order. Copy and paste it.

```{r docker-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Your answer should look like:

```
Layer order matters and mirrors the old base → dev → student layering:
system libs and CLIs first (root), R stacks next (root, then hardened),
then course packages and per-user installs (rstudio), then Python/JS
(root), then the end-to-end smoke tests.
```

###

Docker caches each instruction as a separate layer and reuses cached layers for any instruction that hasn't changed, stopping at the first instruction that *has* changed and rebuilding everything after it. Putting the most stable things first --- system libraries, CLI tools --- means a change deep in the file (say, a new R package) does not force Docker to redo the slow `apt-get` layers. If the order were reversed, almost any edit would invalidate the whole cache and every rebuild would be slow.

### Exercise 4

Find the five `ARG` lines that pin tool versions (`QUARTO_VERSION`, `ARF_VERSION`, `NODE_MAJOR`, `CLAUDE_CODE_VERSION`, `CODEX_VERSION`). Copy and paste them, and note which CLI mentioned nearby is the *exception* to this pinning approach.

```{r docker-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Your answer should look like:

```
ARG QUARTO_VERSION=1.9.38
ARG ARF_VERSION=0.3.4
ARG NODE_MAJOR=22
ARG CLAUDE_CODE_VERSION=2.1.181
ARG CODEX_VERSION=0.141.0
```

The exception is the Antigravity CLI (`agy`) --- its installer offers no version pin, so it always installs whatever is latest.

###

Pinning versions with `ARG` means every build of this Dockerfile installs the *exact same* Quarto, `arf`, Node, Claude Code, and Codex versions until someone deliberately bumps the number --- a reproducible build. `agy` cannot be pinned this way because its installer has no version-pin mechanism, so it floats with each rebuild. That is an accepted, isolated exception: the CLI smoke test near the bottom of the file still fails the build if a future `agy` release breaks.

### Exercise 5

Find the bullet point in the system-libraries comment that explains why `libuv1-dev` is baked into the image even though the R package that needs it doesn't strictly require it. Copy and paste that bullet.

```{r docker-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

Your answer should look like:

```
- libuv1-dev: fs >= 2.1.0 declares libuv as a sysreq. The P3M binary works
   without it, but pak prints a phantom "✖ Missing 1 system package:
   libuv1-dev" on every install that touches fs (i.e. nearly all of them) —
   same annoyance class as the libnode-dev story below, solved here by just
   baking the real (small) lib.
```

###

The `fs` package's P3M binary already works without `libuv1-dev` on the system --- but `pak` still checks the package's declared system requirements and prints an alarming "Missing system package" message every time, even though nothing is actually broken. Since `fs` is a dependency of `primer.tutorials` and gets installed constantly, that phantom warning would appear on nearly every student install. Baking the small library is simpler than trying to suppress the warning.

### Exercise 6

Find the line inside the `libnode-dev-virtual` package's control file that lists what it `Provides`. Copy and paste it, and explain in one sentence why this trick exists.

```{r docker-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look like:

```
'Provides: libnode-dev, libv8-dev' \
```

###

`pak`'s sysreqs database says the V8 R package needs the Debian package `libnode-dev`, but the actual V8 binary from Posit Package Manager is statically linked and needs no system library at all --- and installing a real `libnode-dev` would conflict with the NodeSource `nodejs` package installed earlier. Rather than live with a scary but false "Missing system package" warning on every install that touches V8 (like `gt`), the Dockerfile builds a tiny metadata-only `.deb` that *claims* to provide `libnode-dev`/`libv8-dev` without shipping any files, satisfying `pak`'s check without installing anything real.

### Exercise 7

Find the two-part comment (in the "hardenings" block and the "Everything below installs as rstudio" block) that explains why a `pak` run as root can break later installs by the non-root user. Copy and paste the sentence describing what happens to `filelock`.

```{r docker-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

Your answer should look like:

```
Locks created by ROOT here are group-unreadable (root's
umask), and a later rstudio-user pak run then dies with
"filelock::lock(): Cannot open lock file: Permission denied" — this
broke the student build (2026-07).
```

###

`pkgdepends` (used by `pak`) locks package directories while installing. When those locks are created by root, they inherit root's restrictive umask and end up unreadable by other users. If a *later* install runs as the non-root `rstudio` user (as most of the R installs in this Dockerfile do, after `USER rstudio`), it cannot even open the leftover lock file and the build --- or a student's own `pak::pkg_install()` --- fails outright. This is why the site-library's `_cache` directory is deleted and permissions are widened with `chmod -R g+rwX` right after the root-run installs, and why everything after that point runs `USER rstudio` instead of root.

### Exercise 8

Find the `ENV` line that turns off `pak`'s system-requirement auto-install behavior. Copy and paste it.

```{r docker-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

Your answer should look like:

```
ENV PKG_SYSREQS=false
```

###

Without this, `pak` will try to run `sudo apt-get install` whenever it thinks a system library is missing. The non-root `rstudio` user cannot `sudo`, so that attempt just fails --- and, as with the V8/`libnode-dev` case, `pak` sometimes *mis-detects* a library as missing even though it is already baked into the image. Turning this off stops `pak` from trying to install anything at the system level; it will still print any system requirements it finds, but the build (and a student's runtime installs) relies on the libraries already baked in near the top of the file.

### Exercise 9

Find the `RUN` block that installs the four PPBDS course packages from GitHub. Copy and paste it, and note the one argument to `pak::pkg_install()` that is not just the package list.

```{r docker-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 10)
```

###

Your answer should look like:

```
RUN R -q -e 'pak::pkg_install(c( \
        "PPBDS/tutorial.helpers", \
        "PPBDS/vscode.tutorials", \
        "PPBDS/misc.tutorials", \
        "PPBDS/primer.tutorials" \
    ), upgrade = TRUE)'
```

The extra argument is `upgrade = TRUE`.

###

Installing straight from each package's GitHub default branch (rather than CRAN or an r-universe build) matches PPBDS practice: the ecosystem expects the development version. `upgrade = TRUE` forces `pak` to also pull the latest version of every transitive dependency --- `learnr`, `knitr`, `rmarkdown`, `xfun`, and so on. Without it, a binary dependency built against a newer `xfun` than the one Rocker shipped can fail to load with an error like `object 'attr' is not exported by 'namespace:xfun'` --- exactly the kind of ABI mismatch the smoke test just below this block is designed to catch.

### Exercise 10

Find the `remotes::install_url()` line that installs CatBoost. Copy and paste it, and explain in one sentence why CatBoost needs this special install method instead of a normal `pak::pkg_install()` call.

```{r docker-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

Your answer should look like:

```
R -q -e 'remotes::install_url("https://github.com/catboost/catboost/releases/download/v1.2.10/catboost-R-linux-x86_64-1.2.10.tgz", INSTALL_opts = c("--no-multiarch", "--no-test-load", "--no-staged-install"))'
```

###

CatBoost is not on CRAN, so there is no package for `pak` to find --- it has to be installed directly from CatBoost's own pinned release binary for linux-x86_64 (which is fine, since the image only targets amd64). It is installed with `--no-test-load`, meaning the install step itself never confirms the package actually works. That is exactly why "Modeling smoke test 2" near the bottom of the file goes further than every other smoke test here: it doesn't just load the `catboost` package, it fits a real model through `tidymodels`/`bonsai` and checks the predictions, proving the binary and the engine binding both actually work.

### Exercise 11

Find the `ENV` line that sets `RETICULATE_PYTHON`, and the comment just above the Python venv block explaining why it is set even though nothing installed in this Dockerfile uses it directly. Copy and paste the `ENV` line.

```{r docker-11}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

Your answer should look like:

```
ENV PATH=/opt/venv/bin:$PATH \
    RETICULATE_PYTHON=/opt/venv/bin/python
```

###

Quarto renders Python chunks through the Jupyter engine (`ipykernel`), which needs no bridge from R at all --- so nothing baked into this image actually requires `RETICULATE_PYTHON`. It is set anyway as a convenience: *if* a student later installs R's `reticulate` package (not baked in), it will automatically connect to this same `/opt/venv` environment instead of finding, or creating, a separate Python installation. The Python packages themselves are pinned in `requirements.lock`, generated from `requirements.txt` via `uv pip compile` --- and a separate build step fails the build if `requirements.txt` and `requirements.lock` ever fall out of sync, so an edited `.txt` file can never silently ship with a stale lock.

### Exercise 12

Find the line that installs Observable Framework. Copy and paste it, and explain when a student actually needs this CLI versus when Quarto already has them covered.

```{r docker-12}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

Your answer should look like:

```
RUN npm install -g @observablehq/framework@1.13.4 && observable --version
```

###

Interactive Observable charts *inside* an existing Quarto document need nothing extra --- Quarto already bundles support for `{ojs}` cells. The `observable` CLI installed here is only needed for a different use case: building a standalone Observable Framework data app or dashboard project (`observable create`, `observable preview`, `observable build`), which is a separate kind of project from a Quarto document.

### Exercise 13

Find the "Modeling smoke test 3" `RUN` line near the bottom of the file, the one that fits a `brms` model. Copy and paste it, and explain why this test needs to actually fit a model rather than just checking that the `brms` package loads.

```{r docker-13}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 8)
```

###

Your answer should look like:

```
RUN R --vanilla -e 'fit <- brms::brm(mpg ~ wt, data = mtcars, chains = 1, iter = 100, refresh = 0, silent = 2); stopifnot(inherits(fit, "brmsfit")); cat("brms + Stan end-to-end OK\n")'
```

###

`brms` fits Bayesian models by compiling and running a Stan program, which depends on the image's C++ toolchain working correctly --- a problem that a simple `requireNamespace("brms")` load check would never catch, since the package can load fine even if Stan compilation is broken. Fitting a tiny model (one chain, 100 iterations --- just enough to prove compilation and sampling both run, not to produce a good estimate) is what actually proves the whole modeling stack works. This is the same idea behind every other smoke test in this section: catch a broken build *here*, during the image build, rather than the first time a student tries to use the feature.

## Summary
###

This tutorial read through the real Dockerfile behind `ghcr.io/ppbds/devcontainer`, the image `codespace-starter`'s `devcontainer.json` points at. You saw why the base image is pinned by digest rather than a floating tag, why one image now serves both students and package developers, why layer order matters for Docker's build cache, and how deliberate `ARG` version pins keep builds reproducible. You also traced the permissions story that runs through the R package installs --- system libraries baked in as root, `PKG_SYSREQS=false` to stop `pak` from trying to `sudo`, and the switch to `USER rstudio` so lock files never poison the library for later installs --- and saw how the Python venv and the image-wide smoke tests each guard against a specific class of failure at build time, before a student ever opens a Codespace.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
