---
title: "Codespace Starter Devcontainer"
author:
  - David Kane
  - Krrish Jayakumar
tutorial:
  id: "11-codespace-starter"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "A tour of the devcontainer.json behind codespace-starter itself --- the custom PPBDS image, the VS Code extensions array, the settings that control auto-save, plots, and Git commits, and the postCreateCommand that keeps course packages fresh."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In `10-devcontainer` you examined a generic Python devcontainer and wrote a minimal R devcontainer from scratch. Now you will look at the devcontainer behind the Codespace you are working in right now --- the one that controls behaviors you've likely already noticed, like auto-save, plots appearing inside the editor, and the Git commit button working without an extra confirmation step. `codespace-starter` uses a custom image built specifically for this course, plus a detailed `settings` block that configures all of it.

### Exercise 1

Create and connect to a repo called `11-codespace-starter`, then open it as your workspace.

Copy and paste the URL for your `11-codespace-starter` repository.

```{r introduction-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://github.com/ppbds-student/11-codespace-starter
```

###

This new repo is empty except for a `.git` folder --- the `devcontainer.json` that actually configured this Codespace lives somewhere else entirely, which you'll track down in a couple of exercises.

### Exercise 2

Copy and paste the URL for your Codespace.

```{r introduction-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://congenial-space-broccoli-7qpj5xvw9xhpqj9.github.dev/
```
Your answer should be similar to ours.

###

Every Codespace runs on its own remote machine that GitHub provisions for you --- this URL is that machine's unique address, and it stays the same for the rest of this tutorial.

### Exercise 3

In the bash Terminal, run `cd ../codespace-starter`, then `ls .devcontainer`. CP/CR.

```{r introduction-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 4)
```

###

```
$ cd ../codespace-starter
$ ls .devcontainer
AGENTS.md  STUDENT_WORKFLOW.md  connect-repo.sh  devcontainer.json  welcome.sh
```

###

Your new repo and `codespace-starter` sit side by side under `/workspaces` in this same Codespace. `codespace-starter` is the checkout GitHub actually read to build this machine, including the `devcontainer.json` you'll spend the rest of this tutorial reading.

## Codespace Starter Devcontainer
###

This section walks through `codespace-starter’`s `.devcontainer/devcontainer.json` field by field, examining the custom image, VS Code extensions, settings, and `postCreateCommand`. Unlike the minimal configurations from `10-devcontainer`, this setup customizes almost everything to guarantee an identical experience for every student. By the end, you'll understand each top-level field and manually install a package to see exactly what `postCreateCommand` automates versus what you still do by hand.

### Exercise 1

In the bash Terminal, run `cd /workspaces/codespace-starter`, then `cat .devcontainer/devcontainer.json` to see the whole file on your screen. Find the line beginning with `"name"`. Copy and paste it.

```{r codespace-starter-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"name": "PPBDS Codespace",
```

###

The `name` field is purely cosmetic --- it is the label GitHub shows for this configuration in the Codespace-creation picker and the VS Code window title.

### Exercise 2

Find the line beginning with `"image"`. Copy and paste it.

```{r codespace-starter-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"image": "ghcr.io/ppbds/devcontainer:1.0.1",
```

###

Your `codespace-starter` uses a pinned version of a custom image, `ghcr.io/ppbds/devcontainer`, so every student gets the exact same setup, pre-loaded with Quarto, R, Python, the four course packages, and AI tools. The next tutorial, `12-docker`, shows how this image is actually built.

### Exercise 3

Find the `"extensions"` array. Find the line containing `"reditorsupport.r"`. Copy and paste it.

```{r codespace-starter-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"reditorsupport.r",
```

###

`reditorsupport.r` is the extension that gives VS Code R syntax highlighting, code completion, and the R Terminal integration you've been using all course.

### Exercise 4

Find the line containing `"quarto.quarto"`. Copy and paste it.

```{r codespace-starter-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"quarto.quarto",
```

###

`quarto.quarto` is what lets VS Code render and preview Quarto documents directly in the editor.

### Exercise 5

Find the line containing `"PPBDS.vscode-r-tutorials"`. Copy and paste it.

```{r codespace-starter-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"PPBDS.vscode-r-tutorials",
```

###

`PPBDS.vscode-r-tutorials` is the one extension unique to this course --- it's the panel this tutorial itself is running in right now.

### Exercise 6

Find the two lines containing `"ms-python.python"` and `"ms-toolsai.jupyter"`, along with the comment directly above them. Copy and paste all of it.

```{r codespace-starter-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

```
// Python data science: language support + notebooks. Unlike the AI
// assistant panels above, these are ordinary language tooling that
// renders fine in the web Codespaces client.
"ms-python.python",
"ms-toolsai.jupyter"
```

###

Standard JSON doesn't allow comments, but `devcontainer.json` is actually **JSONC** (JSON with Comments) --- that's what lets the `//` note above explain why these two extensions are grouped separately. `ms-python.python` and `ms-toolsai.jupyter` add Python language support and Jupyter notebook support.

### Exercise 7

Find the lines containing `"ritwickdey.LiveServer"`, `"tomoki1207.pdf"`, and `"mechatroner.rainbow-csv"`. Copy and paste all three.

```{r codespace-starter-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
"ritwickdey.LiveServer",
"tomoki1207.pdf",
"mechatroner.rainbow-csv",
```

###

Three quality-of-life extensions: `ritwickdey.LiveServer` previews HTML files, `tomoki1207.pdf` opens PDFs inside the editor, and `mechatroner.rainbow-csv` colors CSV columns to make them easier to read. Notice what's missing from this whole array: no AI-assistant extensions, because those tools run from the terminal instead, where they work on every device and render properly inside a browser-based Codespace.

### Exercise 8

Find the line containing `"r.rterm.linux"`. Copy and paste it.

```{r codespace-starter-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.rterm.linux": "/usr/local/bin/arf",
```

###

`arf` is a faster stand-in for the default R Terminal --- it starts quicker and handles pasted multi-line code more reliably. `/usr/local/bin/arf` is a fixed location the image always installs it at, so this setting keeps working even if the image changes internally.

### Exercise 9

Find the line containing `"r.plot.useHttpgd"`. Copy and paste it.

```{r codespace-starter-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.plot.useHttpgd": true,
```

###

By default, R opens plots in a separate window that would be invisible in a Codespace. Setting this to `true` sends plots to the **httpgd** tool instead, which streams them into a panel inside VS Code.

### Exercise 10

Find the line containing `"r.source.echo"`. Copy and paste it.

```{r codespace-starter-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.source.echo": true,
```

###

The Run Source button (▷ in the editor title bar) normally runs a file silently, showing only the final output. This setting makes it run with `echo = TRUE` instead, so each line prints as it executes.

### Exercise 11

Find the line containing `"git.enableSmartCommit"`. Copy and paste it.

```{r codespace-starter-11}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"git.enableSmartCommit": true,
```

###

Smart commit automatically includes all changed files in the commit when you click the Commit button in the Source Control sidebar --- you do not need to run `git add` separately. This is why clicking Commit in the `codespace-starter` Codespace works without an extra confirmation step, while the Jupyter Notebook devcontainer setup required one.

### Exercise 12

Find the line containing `"files.autoSave"`. Copy and paste it.

```{r codespace-starter-12}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"files.autoSave": "afterDelay",
```

###

With `"afterDelay"`, VS Code automatically saves open files a short time after you stop typing. This prevents the most common beginner mistake in coding environments: running code that still reflects the last-saved version rather than the current edits.

### Exercise 13

Find the line containing `"editor.wordWrap"`. Copy and paste it.

```{r codespace-starter-13}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"editor.wordWrap": "on",
```

###

Word wrap causes long lines to wrap to the next line visually rather than extending off the right edge of the screen. Without it, long comment blocks or URLs in code appear truncated on smaller screens.

### Exercise 14

Find the line containing `"workbench.editor.enablePreview"`. Copy and paste it.

```{r codespace-starter-14}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"workbench.editor.enablePreview": false,
```

###

By default, VS Code opens files in "preview" mode --- clicking a second file replaces the first tab, so beginners often think their file disappeared. Setting this to `false` means every file opens in a permanent tab.

### Exercise 15

Find the line beginning with `"python.defaultInterpreterPath"`. Copy and paste it.

```{r codespace-starter-15}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"python.defaultInterpreterPath": "/opt/venv/bin/python",
```

###

`/opt/venv` is a **virtual environment**: a self-contained folder of Python packages --- numpy, pandas, matplotlib, and more --- kept separate from the rest of the machine. This setting points VS Code's Python extension at that environment, so `import pandas` works immediately with no setup.

### Exercise 16

Find the line beginning with `"postCreateCommand"`. Copy and paste it.

```{r codespace-starter-16}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
"postCreateCommand": "R -q -e 'pak::pkg_install(c(\"PPBDS/tutorial.helpers\", \"PPBDS/vscode.tutorials\", \"PPBDS/misc.tutorials\", \"PPBDS/primer.tutorials\"), ask = FALSE)'",
```

###

This runs automatically when your Codespace opens, refreshing the four course packages to their latest versions even though the image itself stays pinned. It only ever reinstalls these four packages, though --- anything else you need, you install yourself, by hand or by asking AI, which is exactly what the next few exercises do.

### Exercise 17

In the R Terminal, run `install.packages("praise")`. CP/CR.

```{r codespace-starter-17}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 14)
```

###

```
> install.packages("praise")
Installing package into '/home/ppbds-student/R/library'
(as 'lib' is unspecified)
trying URL 'https://cloud.r-project.org/src/contrib/praise_1.0.0.tar.gz'
Content type 'application/x-gzip' length 6431 bytes
downloaded 6431 bytes

* installing *source* package 'praise' ...
** package 'praise' successfully unpacked and MD5 sums checked
** using staged installation
** R
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** testing if installed package can be loaded from temporary location
** testing if installed package can be loaded from final location
** DONE (praise)
```

###

`install.packages()` is base R's package manager: it downloads one package from CRAN and compiles or unpacks it into your library, one call at a time. Contrast this with **pak** (Exercise 2 and 16): pak resolves and installs entire dependency trees in parallel, which is why `postCreateCommand` uses `pak::pkg_install()` for the four course packages.

### Exercise 18

In the bash Terminal, run `/opt/venv/bin/pip install emoji`. CP/CR.

```{r codespace-starter-18}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

```
$ /opt/venv/bin/pip install emoji
Collecting emoji
  Downloading emoji-2.14.0-py3-none-any.whl (590 kB)
Installing collected packages: emoji
Successfully installed emoji-2.14.0
```

###

That's the same `/opt/venv` from Exercise 15. Calling `/opt/venv/bin/pip` directly, instead of just `pip`, guarantees the package lands in the same environment `python.defaultInterpreterPath` points VS Code at, so `import emoji` works immediately with no restart.

### Exercise 19

Ask AI to install the R package `cowsay` for you, and to tell you exactly what command it ran. CP/CR.

```{r codespace-starter-19}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```
> install.packages("cowsay")
Installing package into '/home/ppbds-student/R/library'
(as 'lib' is unspecified)
trying URL 'https://cloud.r-project.org/src/contrib/cowsay_0.9.0.tar.gz'
downloaded 22 KB

* installing *source* package 'cowsay' ...
** package 'cowsay' successfully unpacked and MD5 sums checked
** using staged installation
** R
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
** building package indices
** DONE (cowsay)
```

###

Whether you type `install.packages()` yourself or ask AI to install cowsay, the same function ultimately runs underneath --- an AI agent calls the same R functions and Terminal commands you would type by hand. That's why it's worth reading the exact command AI reports back, the same way you read your own in Exercise 17.

### Exercise 20

Ask AI to install the Python library `pyfiglet` into this Codespace's Python environment, and to tell you exactly what command it ran. CP/CR.

```{r codespace-starter-20}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 6)
```

###

```
$ /opt/venv/bin/pip install pyfiglet
Collecting pyfiglet
  Downloading pyfiglet-1.0.2-py3-none-any.whl (1.1 MB)
Installing collected packages: pyfiglet
Successfully installed pyfiglet-1.0.2
```

###

A good AI agent in this Codespace targets `/opt/venv/bin/pip` rather than plain `pip`, because that's the interpreter `python.defaultInterpreterPath` (Exercise 15) points VS Code at --- an assistant without that context might install into the wrong environment and leave `import pyfiglet` failing silently later. Checking which `pip` actually ran is worth doing before trusting the result, not after.

## Summary
###

This tutorial toured the `devcontainer.json` behind `codespace-starter` itself: its pinned custom PPBDS image, its VS Code extensions array, the settings that control the R Terminal, plot rendering, Git commits, autosave, word wrap, tab behavior, and the Python interpreter, and the `postCreateCommand` that keeps the four fast-moving course packages current. The next tutorial, `12-docker`, goes one level deeper and reads through the Dockerfile that actually builds the `ghcr.io/ppbds/devcontainer` image this `devcontainer.json` points at.

### Exercise 1

In the bash Terminal, in your `11-codespace-starter` repo, run `git status`. CP/CR.

```{r summary-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
11-codespace-starter $ git status
On branch main
nothing to commit, working tree clean
11-codespace-starter $
```

###

A clean `git status` here just confirms Git isn't tracking anything you haven't dealt with --- this tutorial never asked you to change files in your own repo, only to read `codespace-starter`'s.

### Exercise 2

Paste the URL of your `11-codespace-starter` repository.

```{r summary-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
https://github.com/ppbds-student/11-codespace-starter
```

###

This repo now exists on GitHub independent of any one Codespace --- you or a grader can reconnect to it and pick your work back up even after the Codespace itself is gone.

### Exercise 3

Stop this Codespace: open the Command Palette with `Ctrl+Shift+P` (or `Cmd+Shift+P` on Mac), type `>stop`, and select **"Codespaces: Stop Current Codespace"**. Then go to https://github.com/codespaces and paste the row for it, showing its "Stopped" status.

```{r summary-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
11-codespace-starter    ppbds-student/11-codespace-starter    main    Stopped
```

###

Stopped Codespaces don't use compute hours but still occupy storage, so they can be restarted any time --- deleting one instead frees the storage but can't be undone.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
