---
title: "Codespace Starter Devcontainer"
author:
  - David Kane
  - Krrish Jayakumar
tutorial:
  id: "11-codespace-starter"
output:
  learnr::tutorial:
    df_print: default
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "A tour of the devcontainer.json behind codespace-starter itself --- the custom PPBDS image, the VS Code extensions array, the settings that control auto-save, plots, and Git commits, and the postCreateCommand that keeps course packages fresh."
---

```{r setup, include = FALSE}
library(learnr)
library(tutorial.helpers)
library(knitr)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(out.width = '90%')
options(tutorial.exercise.timelimit = 60,
        tutorial.storage = "local")
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

## Introduction
###

In `10-devcontainer` you examined a generic Python devcontainer and wrote a minimal R devcontainer from scratch. Now you will look at the devcontainer behind the Codespace you are working in right now. `codespace-starter` uses a custom image built specifically for this course and a detailed `settings` block that configures VS Code's behavior. Many of the behaviors you may have noticed --- auto-save, plots appearing inside the editor, the Git commit button working without an extra confirmation step --- are controlled here.

###

You do not need to create a new repo for this tutorial. All of the exercises below work directly inside the `codespace-starter` checkout that is already part of this Codespace.

## Codespace Starter Devcontainer
###

### Exercise 1

In the bash Terminal, run `cd /workspaces/codespace-starter`, then `cat .devcontainer/devcontainer.json`. Copy and paste the entire output.

```{r codespace-starter-1}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 30)
```

###

````
{
  "name": "PPBDS Codespace",
  "image": "ghcr.io/ppbds/devcontainer:1.0.1",
  "customizations": {
    "vscode": {
      "extensions": [
        "reditorsupport.r",
        "quarto.quarto",
        "PPBDS.vscode-r-tutorials",
        "ritwickdey.LiveServer",
        "tomoki1207.pdf",
        "mechatroner.rainbow-csv",
        // Python data science: language support + notebooks. Unlike the AI
        // assistant panels above, these are ordinary language tooling that
        // renders fine in the web Codespaces client.
        "ms-python.python",
        "ms-toolsai.jupyter"
      ],
      "settings": {
        "r.rterm.linux": "/usr/local/bin/arf",
        "r.plot.useHttpgd": true,
        "r.source.echo": true,
        "git.enableSmartCommit": true,
        "files.autoSave": "afterDelay",
        "editor.wordWrap": "on",
        "workbench.editor.enablePreview": false,
        "python.defaultInterpreterPath": "/opt/venv/bin/python"
      }
    }
  },
  "postCreateCommand": "R -q -e 'pak::pkg_install(c(\"PPBDS/tutorial.helpers\", \"PPBDS/vscode.tutorials\", \"PPBDS/misc.tutorials\", \"PPBDS/primer.tutorials\"), ask = FALSE)'"
}
````

Note that the exact `name` field, key order, and image version may differ slightly on your machine as the image is updated.

###

Standard JSON doesn't allow comments, but `devcontainer.json` uses **JSONC** (JSON with Comments), which does. The comments here explain *why* each setting exists --- useful for instructors and future maintainers.

### Exercise 2

Find the line beginning with `"image"`. Copy and paste it.

```{r codespace-starter-2}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"image": "ghcr.io/ppbds/devcontainer:1.0.1",
```

Note that the version number will change as the image is updated.

###

Your `codespace-starter` uses a pinned version of a custom image, `ghcr.io/ppbds/devcontainer`, so every student gets the exact same setup. No surprises, no version conflicts. The image comes pre-loaded with everything you need: `Quarto`, `gh`, the `arf R console`, **pak**, our four course packages (**tutorial.helpers**, **vscode.tutorials**, **misc.tutorials**, **primer.tutorials**), R and Python data-science stacks, and AI assistants for your terminal. Next up, 12-docker shows how this image is actually built.

### Exercise 3

Find the `"extensions"` array. Copy and paste the entire block, from the line containing `"extensions": [` through the closing `]`.

```{r codespace-starter-3}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 12)
```

###

```
"extensions": [
  "reditorsupport.r",
  "quarto.quarto",
  "PPBDS.vscode-r-tutorials",
  "ritwickdey.LiveServer",
  "tomoki1207.pdf",
  "mechatroner.rainbow-csv",
  // Python data science: language support + notebooks. Unlike the AI
  // assistant panels above, these are ordinary language tooling that
  // renders fine in the web Codespaces client.
  "ms-python.python",
  "ms-toolsai.jupyter"
],
```

###

Each extension has one job: `reditorsupport.r` for R, `quarto.quarto` for Quarto, `PPBDS.vscode-r-tutorials` for the tutorial panel, `ritwickdey.LiveServer` for HTML previews, `tomoki1207.pdf` for PDFs, `mechatroner.rainbow-csv` for colored CSVs, and `ms-python.python` plus `ms-toolsai.jupyter` for Python and Jupyter. Notice what's missing: no AI-assistant extensions. Instead, AI tools run from the terminal—they work better as CLI programs on every device, and in-editor panels don't render well in browser-based Codespaces.

### Exercise 4

Find the line containing `"r.rterm.linux"`. Copy and paste it.

```{r codespace-starter-4}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.rterm.linux": "/usr/local/bin/arf",
```

###

`arf` is a fast, Rust-based stand-in for the default R Terminal --- it starts quicker and handles pasted multi-line code more reliably. The path `/usr/local/bin/arf` is a stable symlink the image sets up just for this, so the setting keeps working even if the image's internals change.

### Exercise 5

Find the line containing `"r.plot.useHttpgd"`. Copy and paste it.

```{r codespace-starter-5}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.plot.useHttpgd": true,
```

###

By default, R opens plots in a separate window that would be invisible in a Codespace. Setting this to `true` sends plots to the **httpgd** tool instead, which streams them into a panel inside VS Code.

### Exercise 6

Find the line containing `"r.source.echo"`. Copy and paste it.

```{r codespace-starter-6}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"r.source.echo": true,
```

###

The "Run Source" button (the ▷ in the editor title bar) and its `Ctrl/Cmd + Shift + S` shortcut normally run a file silently, showing only the final output. This setting makes them run `source()` with `echo = TRUE` instead, so each line of code is printed as it executes --- when something goes wrong, you can see exactly which line produced the error.

### Exercise 7

Find the line containing `"git.enableSmartCommit"`. Copy and paste it.

```{r codespace-starter-7}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"git.enableSmartCommit": true,
```

###

Smart commit automatically includes all changed files in the commit when you click the Commit button in the Source Control sidebar --- you do not need to run `git add` separately. This is why clicking Commit in the `codespace-starter` Codespace works without an extra confirmation step, while the Jupyter Notebook devcontainer setup required one.

### Exercise 8

Find the line containing `"files.autoSave"`. Copy and paste it.

```{r codespace-starter-8}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"files.autoSave": "afterDelay",
```

###

With `"afterDelay"`, VS Code automatically saves open files a short time after you stop typing. This prevents the most common beginner mistake in coding environments: running code that still reflects the last-saved version rather than the current edits, because the file was never manually saved.

### Exercise 9

Find the line containing `"editor.wordWrap"`. Copy and paste it.

```{r codespace-starter-9}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"editor.wordWrap": "on",
```

###

Word wrap causes long lines to wrap to the next line visually rather than extending off the right edge of the screen. Without it, long comment blocks or URLs in code appear truncated, and students on small screens may not notice that content is hidden to the right.

### Exercise 10

Find the line containing `"workbench.editor.enablePreview"`. Copy and paste it.

```{r codespace-starter-10}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"workbench.editor.enablePreview": false,
```

###

By default, VS Code opens files in "preview" mode --- clicking a second file replaces the first tab, so beginners often think their file disappeared. Setting this to `false` means every file opens in a permanent tab.

### Exercise 11

Find the line beginning with `"python.defaultInterpreterPath"`. Copy and paste it.

```{r codespace-starter-11}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 3)
```

###

```
"python.defaultInterpreterPath": "/opt/venv/bin/python",
```

###

The image ships a full Python data-science stack --- **numpy**, **pandas**, **matplotlib**, **scikit-learn**, and more --- in a virtual environment at `/opt/venv`. This setting points the Python extension at that environment, so `import pandas` works in a Jupyter notebook or a Quarto Python chunk with no setup. This is a recent addition: earlier versions of `codespace-starter` were R-only.

### Exercise 12

Find the line beginning with `"postCreateCommand"`. Copy and paste it.

```{r codespace-starter-12}
question_text(NULL,
	answer(NULL, correct = TRUE),
	allow_retry = TRUE,
	try_again_button = "Edit Answer",
	incorrect = NULL,
	rows = 5)
```

###

```
"postCreateCommand": "R -q -e 'pak::pkg_install(c(\"PPBDS/tutorial.helpers\", \"PPBDS/vscode.tutorials\", \"PPBDS/misc.tutorials\", \"PPBDS/primer.tutorials\"), ask = FALSE)'",
```

###

This runs automatically when your Codespace opens, refreshing the four course packages — **tutorial.helpers**, **vscode.tutorials**, **misc.tutorials**, **primer.tutorials** — to their latest versions. You always get the newest tutorials even though the image is pinned. A working copy is baked into the image as a fallback, so GitHub outages won't block your Codespace from opening.

## Summary
###

This tutorial toured the `devcontainer.json` behind `codespace-starter` itself: its pinned custom PPBDS image, its VS Code extensions array, the settings that control the R Terminal, plot rendering, Git commits, autosave, word wrap, tab behavior, and the Python interpreter, and the `postCreateCommand` that keeps the four fast-moving course packages current. The next tutorial, `12-docker`, goes one level deeper and reads through the Dockerfile that actually builds the `ghcr.io/ppbds/devcontainer` image this `devcontainer.json` points at.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```
