---
title: "Your Starter"
author: David Kane
description: "Graduation exercise for the devcontainer tutorials. Build your own working version of codespace-starter: a repo named my-starter with your own Dockerfile, an image built and published by a GitHub Actions workflow, and a devcontainer.json that pins it --- then launch a Codespace from it and prove it works."
format: live-html
engine: knitr
toc: true
---

{{< include _extensions/r-wasm/live/_knitr.qmd >}}

## Information
###

```{r}
#| label: information-1
#| echo: false
learnr2::student_info()
```

## Introduction
###

In the last three tutorials you read a real `devcontainer.json`, a real Dockerfile, and the workflow that turns that Dockerfile into a published image. Now you build the whole chain yourself, in miniature: a repo called `my-starter` that you can launch a Codespace from --- your own `codespace-starter`, without the teaching material you no longer need.

"Miniature" means one base image, one extra R package, one extension, and three settings. But every piece is real, and the shape is identical to the PPBDS version: a Dockerfile describes the image, a GitHub Actions workflow builds and publishes it, and a `devcontainer.json` pins it by exact version.

Two rules of the road. First, you never run `docker` yourself; a machine does, exactly as in the previous tutorial. Second, you keep this repo. Every other tutorial repo was disposable. This one is a tool you will use.

### Exercise 1

You should be doing this tutorial in a repo named `my-starter`. If you are not, create one and then connect to it. You may need to restart this tutorial after you do so.

Copy and paste the URL for your `my-starter` repository.

```{r}
#| label: introduction-1
#| echo: false
learnr2::question(
  "Copy and paste the URL for your my-starter repository.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
https://github.com/ppbds-student/my-starter
```

###

That repo is the product. Everything you make in this tutorial --- the Dockerfile, the workflow, the `devcontainer.json` --- lives inside it, and at the end you will turn it into a **template** so new projects can start from it with one click.

## Your Starter
###

Build from the bottom up, in the order the pieces depend on each other: the Dockerfile first, then the workflow that builds it, then the `devcontainer.json` that uses the result.

### Exercise 1

In the bash Terminal, at the top level of your `my-starter` repo, create a file called `Dockerfile` (no extension) with exactly this content:

```
FROM ghcr.io/rocker-org/devcontainer/tidyverse:4.5

# One package of your own, so this is provably YOUR image.
RUN R -q -e 'install.packages("praise")'

# Smoke test: fail the build if the package did not install.
RUN R -q -e 'stopifnot(requireNamespace("praise"))'
```

Then run `cat Dockerfile`. CP/CR.

```{r}
#| label: your-starter-1
#| echo: false
learnr2::question(
  "Then run cat Dockerfile. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ cat Dockerfile
FROM ghcr.io/rocker-org/devcontainer/tidyverse:4.5

# One package of your own, so this is provably YOUR image.
RUN R -q -e 'install.packages("praise")'

# Smoke test: fail the build if the package did not install.
RUN R -q -e 'stopifnot(requireNamespace("praise"))'
```

###

Three lines of real content. The `FROM` line starts from the same Rocker family as the PPBDS image --- but at `:4.5`, not `:4.6`. That is deliberate: R 4.6 changed how R starts up in a way that breaks the R extension's plot pane unless the image patches around it. The PPBDS Dockerfile carries exactly such a patch (the `.First` shim you can find near its arf section); you sidestep the whole problem by starting one version back. Pinning by tag rather than by digest is fine for a first version --- the digest is the professional upgrade you read about in Exercise 1 of the previous tutorial.

The two `RUN` lines are the PPBDS discipline in miniature: install something, then **test** that it installed, so a broken build fails loudly instead of shipping a broken image.

### Exercise 2

Now the workflow. Create the folders and file `.github/workflows/build.yml` with this content, **replacing `YOUR_USERNAME` with your GitHub username in lowercase** (registry names must be lowercase):

```
name: build-and-push

on:
  push:
    branches: [main]
    tags: ['v*']

jobs:
  build:
    runs-on: ubuntu-latest
    permissions:
      contents: read
      packages: write
    steps:
      - uses: actions/checkout@v5
      - uses: docker/setup-buildx-action@v4
      - uses: docker/login-action@v4
        with:
          registry: ghcr.io
          username: ${{ github.actor }}
          password: ${{ secrets.GITHUB_TOKEN }}
      - id: meta
        uses: docker/metadata-action@v6
        with:
          images: ghcr.io/YOUR_USERNAME/my-starter
          tags: |
            type=raw,value=latest,enable={{is_default_branch}}
            type=semver,pattern={{version}}
      - uses: docker/build-push-action@v7
        with:
          context: .
          push: true
          tags: ${{ steps.meta.outputs.tags }}
```

Run `cat .github/workflows/build.yml`. CP/CR.

```{r}
#| label: your-starter-2
#| echo: false
learnr2::question(
  "Run cat .github/workflows/build.yml. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ cat .github/workflows/build.yml
name: build-and-push

on:
  push:
    branches: [main]
    tags: ['v*']
```

###

This is a shortened copy of the workflow you read in the previous tutorial --- the same actions at the same versions, minus the caching. Nothing here is new: it checks out your repo, logs in to GHCR with the temporary `GITHUB_TOKEN` (that is what `permissions: packages: write` allows), computes the image tags, then builds and pushes. Two triggers: every push to `main` refreshes `latest`, and a tag like `v1.0.0` publishes `1.0.0` --- which is the one your `devcontainer.json` will pin.

### Exercise 3

Commit and push. Then, in the bash Terminal, run `gh run list --limit 1`. CP/CR.

```{r}
#| label: your-starter-3
#| echo: false
learnr2::question(
  "Commit and push. Then, in the bash Terminal, run gh run list --limit 1. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ gh run list --limit 1
*  Add Dockerfile and build workflow  build-and-push  main  push  31812345678  1m2s  1m
```

###

The `*` means it is still running. The first build is slow --- five minutes or so --- because the runner has to download the ~4 GB Rocker base image before it can add your one package. Re-run the command until the leading symbol turns into a `✓`. If it ever turns into an `X`, click through to the Actions tab on github.com: the log names the exact step that failed, and the cause is almost always a typo in one of the two files you just wrote.

### Exercise 4

Once the build is green, publish a versioned image by creating a release. In the bash Terminal, run `gh release create v1.0.0 --title "v1.0.0" --notes "First image."`. CP/CR.

```{r}
#| label: your-starter-4
#| echo: false
learnr2::question(
  "Once the build is green, publish a versioned image by creating a release. In the bash Terminal, run gh release create v1.0.0 --title \"v1.0.0\" --notes \"First image.\". CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ gh release create v1.0.0 --title "v1.0.0" --notes "First image."
https://github.com/ppbds-student/my-starter/releases/tag/v1.0.0
```

###

Creating the release pushes a git tag `v1.0.0`, which matches the workflow's `tags: ['v*']` trigger, which starts a second build --- and that build's `type=semver` rule names the result `1.0.0`. This is the exact release dance PPBDS performs for every image version. `gh run list --limit 2` will show both builds; wait for the second `✓`.

### Exercise 5

Images pushed by a workflow start out **private**, and a Codespace pulls its image before you are logged in to anything, so a private image would fail. Make yours public. On github.com, open your profile, click **Packages**, open `my-starter`, then **Package settings** (right side), scroll to **Danger Zone**, and **Change visibility** to Public.

Copy and paste the URL of the package page.

```{r}
#| label: your-starter-5
#| echo: false
learnr2::question(
  "Copy and paste the URL of the package page.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
https://github.com/ppbds-student/my-starter/pkgs/container/my-starter
```

###

While you are on that page, notice the tags listed for the package: `latest` from the push build and `1.0.0` from the release build. That is the tag family you read about in the previous tutorial, now with your name on it.

### Exercise 6

Now the file that ties it together. Create `.devcontainer/devcontainer.json` with this content, again **replacing `YOUR_USERNAME` with your lowercase GitHub username**:

```
{
  "name": "my-starter",
  "image": "ghcr.io/YOUR_USERNAME/my-starter:1.0.0",
  "customizations": {
    "vscode": {
      "extensions": [
        "reditorsupport.r@2.8.8"
      ],
      "settings": {
        "r.plot.useHttpgd": true,
        "files.autoSave": "afterDelay",
        "editor.wordWrap": "on"
      }
    }
  },
  "postAttachCommand": "echo 'my-starter is ready'"
}
```

Run `cat .devcontainer/devcontainer.json`. CP/CR.

```{r}
#| label: your-starter-6
#| echo: false
learnr2::question(
  "Run cat .devcontainer/devcontainer.json. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ cat .devcontainer/devcontainer.json
{
  "name": "my-starter",
  "image": "ghcr.io/ppbds-student/my-starter:1.0.0",
```

###

Every field maps to something you read in the Our Codespace Starter tutorial. `"image"` pins **your** image at an exact version, never `latest`, for the reason that tutorial hammered on: a floating tag would change your environment without you deciding it should. The R extension is pinned at `@2.8.8` for the same reason PPBDS pins it. `r.plot.useHttpgd` works because the Rocker base already includes `httpgd`; `files.autoSave` and `editor.wordWrap` are the two beginner-protection settings. And `postAttachCommand` is the simplest possible ready signal --- a one-line `echo` in place of `welcome.sh`.

### Exercise 7

Commit and push. Then run `git log --oneline -n 3`. CP/CR.

```{r}
#| label: your-starter-7
#| echo: false
learnr2::question(
  "Commit and push. Then run git log --oneline -n 3. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ git log --oneline -n 3
a1b2c3d Add devcontainer.json
e4f5a6b Add Dockerfile and build workflow
0f1e2d3 Initial commit
```

###

Three commits, three layers of the same idea: what the image contains, how it gets built, and how a Codespace uses it. Notice that this push triggered one more `latest` build --- harmless, since `devcontainer.json` is not part of the image and nothing in the Dockerfile changed.

### Exercise 8

Time to launch it. On github.com, open your `my-starter` repo, click the green **Code** button, choose the **Codespaces** tab, and click **Create codespace on main**. It will take several minutes: your repo has no prebuild, so the container is built from scratch, pulling your image from GHCR. When the terminal appears, a line should print. Copy and paste it.

```{r}
#| label: your-starter-8
#| echo: false
learnr2::question(
  "Time to launch it. On github.com, open your my-starter repo, click the green Code button, choose the Codespaces tab, and click Create codespace on main. It will take several minutes: your repo has no prebuild, so the container is built from scratch, pulling your image from GHCR. When the terminal appears, a line should print. Copy and paste it.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter is ready
```

###

That line came from your `postAttachCommand`, which means the whole chain held: GitHub read your `devcontainer.json`, pulled `ghcr.io/…/my-starter:1.0.0` from the registry, started it, attached the editor, and ran your command. The slow start is what `codespace-starter` avoids with a prebuild (Settings → Codespaces on github.com); it costs storage, so leave it off until you use this starter often enough to care.

### Exercise 9

In your new Codespace, open an R Terminal (**Cmd/Ctrl + Shift + P**, then "R: Create R Terminal"). Run `library(praise)` and then `praise()`. CP/CR.

```{r}
#| label: your-starter-9
#| echo: false
learnr2::question(
  "In your new Codespace, open an R Terminal (Cmd/Ctrl + Shift + P, then \"R: Create R Terminal\"). Run library(praise) and then praise(). CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
> library(praise)
> praise()
[1] "You are wonderful!"
```

###

`praise` is not part of any Rocker image. It exists in this Codespace for exactly one reason: your Dockerfile installed it. This is the proof that you are running **your** image, not one that happened to be lying around.

### Exercise 10

Still in the new Codespace, in its bash Terminal, run `R --version | head -n 1`. CP/CR.

```{r}
#| label: your-starter-10
#| echo: false
learnr2::question(
  "Still in the new Codespace, in its bash Terminal, run R --version | head -n 1. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ R --version | head -n 1
R version 4.5.3 (2026-03-11) -- "Reassured Reassurer"
```

###

R 4.5, as you chose in the `FROM` line. When you are ready to move to 4.6, you now know exactly what has to change and why: bump the tag, add the startup shim the PPBDS Dockerfile uses, rebuild, release `v1.1.0`, repin. Same loop, one version at a time.

### Exercise 11

Back in your original Codespace, make `my-starter` a template. On github.com, open the repo's **Settings**, and near the top of **General**, check **Template repository**. Then, in the bash Terminal, run `gh repo view my-starter --json isTemplate`. CP/CR.

```{r}
#| label: your-starter-11
#| echo: false
learnr2::question(
  "Back in your original Codespace, make my-starter a template. On github.com, open the repo's Settings, and near the top of General, check Template repository. Then, in the bash Terminal, run gh repo view my-starter --json isTemplate. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ gh repo view my-starter --json isTemplate
{"isTemplate":true}
```

###

This is how you will actually use it. A template repo gains a green **Use this template** button; each project you start that way gets its own repo, owned by you, that inherits `.devcontainer/` --- so launching a Codespace from any of them gives you your environment. That also retires `connect-repo`: that script exists only because students launch from a repo they do not own. Your projects are yours from the first commit.

### Exercise 12

Run `gh release list`. CP/CR.

```{r}
#| label: your-starter-12
#| echo: false
learnr2::question(
  "Run gh release list. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ gh release list
TITLE   TYPE    TAG NAME  PUBLISHED
v1.0.0  Latest  v1.0.0    about 20 minutes ago
```

###

The release is the durable record of what you published, and the loop for every future change runs through it: edit the Dockerfile, push, `gh release create v1.0.1`, wait for the build, change the `1.0.0` in your `devcontainer.json` to `1.0.1`, push again. One last piece of housekeeping: the test Codespace you launched in Exercise 8 is still running. Go to https://github.com/codespaces, find the one on `my-starter`, and stop or delete it --- you will start fresh ones from your template projects rather than reusing this test.

## Summary
###

You built the entire chain you spent three tutorials reading: a Dockerfile that adds one package to a pinned base and tests it; a GitHub Actions workflow that builds and publishes the image on every push and every `v*` release, tagged by pattern, logged in with a token no human handles; a `devcontainer.json` that pins the published image at an exact version and configures the editor; and a Codespace launched from all of it, with your package inside. Turning the repo into a template makes it your `codespace-starter` from here on. To change the environment, edit the Dockerfile, push, release `v1.0.1`, and repin --- the same loop PPBDS runs for every version you have used in this course.

### Exercise 1

In the bash Terminal, in your `my-starter` repo, run `git status`. CP/CR.

```{r}
#| label: summary-1
#| echo: false
learnr2::question(
  "In the bash Terminal, in your my-starter repo, run git status. CP/CR.",
  type = "reflection",
  show_text = FALSE
)
```

###

```
my-starter $ git status
On branch main
Your branch is up to date with 'origin/main'.

nothing to commit, working tree clean
```

###

Clean, pushed, and reusable. That is what a starter is.

## Download answers

```{r}
#| label: download-answers-1
#| echo: false
learnr2::question(
  "How many minutes, approximately, did it take you to complete this
  tutorial? For example, an hour and a half would be 90 minutes.",
  type = "reflection_editable",
  validate = "integer"
)
```

###

```{r}
#| label: download-answers-2
#| echo: false
learnr2::download_answers_button(filename_prefix = "16-your-starter")
```
